---
title: "EDAM training - cleaning and exporing data (solutions)"
author: "Tristan Naidoo"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
    number_sections: false
    theme: flatly
    highlight: tango
    code_folding: show
    df_print: paged
    fig_width: 8
    fig_height: 5
---

> ⚠️ **This is simulated data, for teaching purposes only.**
> All screening, population, and health post data is simulated.
> Nothing here describes a real person, health facility, programme or nutrition outcome.
> Only the rainfall data is real, sourced from [OCHA HDX](https://data.humdata.org/dataset/eth-rainfall-subnational).

## Overview

In this tutorial, imagine you work in an analysis team in a nutrition
office in the Ethiopian Ministry of Health. You have just received a
request from the Oromia Regional Health Bureau. Rainfall has been poor
across the region for the past two seasons, and they want to know whether
this is showing up in child nutrition. Based on their request, you
determine that there are two main questions to answer:

1. **Which zones in Oromia have the highest malnutrition?**
2. **Is rainfall associated with child malnutrition in Oromia?**

To answer these you have growth-monitoring records for children under 5 from
health posts across 56 woredas (districts) in 14 zones, and rainfall estimates
for the same zones and months. There is no budget or time for a survey.

You will work towards an answer in three parts, each building on the
output of the last:

1. **Checking data quality.** You apply six data quality dimensions to the raw
   register: accuracy, completeness, uniqueness, consistency, timeliness, and
   validity. This gives you one documented cleaning pipeline.
2. **Working with shapefiles.** This part focuses on shapefiles and, more
   specifically, on how you get the cleaned register onto a map. It covers
   coordinate reference systems, spatial joins, and the difference between
   dissolving geometry and aggregating data.
3. **Creating an outcome variable and exploring it.** You determine how many
   children are underweight, and convert these to proportions so you can
   compare across zones. You then use exploratory analysis to inspect these
   proportions and answer both questions above.

By the end you will have a clean, geographically located, outcome-scored
dataset, and an answer to both questions along with its limitations.

## Setup
```{r, echo=FALSE}
knitr::opts_chunk$set(fig.align = "center")
```

```{r setup, message=FALSE, warning=FALSE}
# Setup chunk
# Libraries used in this tutorial
library(anthro)
library(dplyr)
library(forcats)
library(ggplot2)
library(lubridate)
library(patchwork)
library(readr)
library(sf)
library(stringr)
library(tidyr)

# Caption for plots
sim_caption <- "*Simulated data - for teaching purposes only"
```

## Part 1 - check your ingredients

Throughout this tutorial you will use four data tables and two shapefiles.
You read in each file as it is required, and a brief explanation of the
dataset and its codebook comes before each one.

In this part, you will focus on nutrition screening as an illustrative example, but in
practice you should check all your files.

#### `nutrition_screening.csv` - one row per child per visit

The raw growth-monitoring register. This captures every visit a child made to a
health post, typed in by hand.

| Field | Type | Description |
|---|---|---|
| `record_id` | text | Unique visit identifier |
| `person_id` | text | Unique person identifier (identifies the person across visits) |
| `woreda_name` | text | Woreda name as typed at the health post |
| `health_post` | text | Unique identifier of the health post the child attended |
| `visit_date` | date | Date of the screening visit |
| `entry_date` | date | Date the record was entered into the system. Compare to `visit_date` for reporting lag |
| `sex` | text | The sex of the child |
| `dob` | date | Date of birth of the child |
| `age_value` | number | Age as recorded, in the unit given by `age_unit` |
| `age_unit` | text | The age unit (`"months"`/`"years"`) |
| `weight_kg` | number | Child's weight in kilograms (kg) |
| `height_cm` | number | Child's height/length in centimetres (cm) |
| `muac_mm` | number | Mid-upper arm circumference (MUAC), in millimetres (mm) |
```{r screening}
screening <- read_csv("data/input/nutrition_screening.csv")
```

### What have you got?

You start by looking at the first few rows, and then at a summary of every
column.

```{r shape}
head(screening)
summary(screening)
```

Three things are already visible:

* **`weight_kg` maximum in the thousands** - grams in a kilogram field.
* **`height_cm` minimum well under 1** - heights entered in metres.
* **`height_cm`, `muac_mm`, `sex`, and `weight_kg` all have missing values.**
  That is check 1 - how much, and does it cluster anywhere?

Every check below works on `cohort_clean` - a single object you build up as you
go, one dimension at a time. Where a dimension has a known, explainable fix,
you apply it immediately and `cohort_clean` carries the fix into the next
check. At the end you collect the same set of decisions into one documented
pipeline, so you can see both ways of getting there.

```{r cohort-init}
cohort_clean <- screening
```

---

### 1. Completeness

To assess completeness, you check what percentage of values are missing per
column and look for patterns.

```{r completeness-a}
cohort_clean |>
  summarise(across(everything(), ~ round(mean(is.na(.)) * 100, 1))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "pct_missing") |>
  arrange(desc(pct_missing))
```

You raise `weight_kg`, `height_cm`, `sex` and `muac_mm` with the regional bureau.
They come back having already looked into three of the four: `weight_kg` and
`height_cm` are **missing completely at random (MCAR)**, meaning the missingness
is unrelated to the data. There is nothing to explain, and nothing to fix beyond
dropping those rows. However, `sex` is missing at the visit level rather than the
child level - a child's other visits usually carry a usable value. These rows
are recovered when we consider consistency.

```{r completeness-drop}
cohort_clean <- cohort_clean |>
  filter(!is.na(weight_kg), !is.na(height_cm))

nrow(cohort_clean)
```

They are less sure about `muac_mm`, though. Their guess is **missing at random
(MAR)**, meaning it is missing for a reason tied to something you can see in
the data. They suggest looking at the health post, since that is the level at
which equipment and staffing actually vary:

```{r completeness-b}
cohort_clean |>
  group_by(health_post) |>
  summarise(muac_missing = round(100 * mean(is.na(muac_mm))), .groups = "drop") |>
  filter(muac_missing > 30) |>
  arrange(desc(muac_missing))
```

Two health posts stand out, one missing 80% of its MUAC readings and
another 58%, while every other post has none at all. That confirms the
bureau's guess: the missingness lines up with `health_post`, a column you can
see, not with the MUAC value itself. This is missing at random (MAR). Most
likely, neither post has a working MUAC tape.

`muac_mm` is not used anywhere else in this analysis, so you do not repair this
missingness.

**Since you take this to be MAR, how would you address the missingness?** The
missingness is explained by an observed column, so condition on that column. If
you needed `muac_mm` downstream, you would impute it using `health_post` as a
predictor rather than a single overall mean that ignores which posts are
measuring at all. Here it is not used downstream, so leave it missing and
document why.

**What would you need to see instead for `muac_mm` to be MNAR? Give a concrete
example.** The missingness would have to depend on the MUAC value itself. For
example, fieldworkers skipping the tape on children who looked visibly thin,
perhaps because those children were fast-tracked into treatment. The lowest
readings would then be the ones missing. Comparing `weight_kg` between children
with and without a reading might hint at this, but a gap fully explained by
`weight_kg` would then become MAR, since `weight_kg` is observed. MNAR means the
gap survives conditioning on every column you have, which is why you can never
confirm it from the data alone.

---

### 2. Accuracy

To assess accuracy, you check whether recorded values are physically possible,
and wherever you know the actual mechanism behind the error, you repair it
rather than remove it.

You contact the regional office that provided this data for guidance. They
confirm three known issues: some weights were mistakenly recorded in grams
rather than kilograms, some heights in metres rather than centimetres, and some
ages in months have been recorded as years. You count how many rows each issue
affects.

```{r accuracy-diagnose}
cohort_clean |>
  summarise(
    weight_bad = sum(weight_kg > 30, na.rm = TRUE),
    height_bad = sum(height_cm < 40, na.rm = TRUE),
    age_bad = sum(age_unit == "years" & age_value > 5, na.rm = TRUE)
  )
```

You then look at a few of the affected rows.

```{r accuracy-values}
cohort_clean |>
  filter(weight_kg > 30 | height_cm < 40 | (age_unit == "years" & age_value > 5)) |>
  select(age_value, age_unit, weight_kg, height_cm) |>
  head(8)
```

Given that you know the cause, you can fix the problematic rows. If you did not
know the cause, you would need to decide whether to drop the rows (assuming
there is no pattern to which ones are problematic) or contact the bureau to
figure out the cause.

```{r accuracy-fix}
cohort_clean <- cohort_clean |>
  mutate(
    age_unit = if_else(age_unit == "years" & age_value > 5, "months", age_unit),
    weight_kg = if_else(weight_kg > 30, weight_kg / 1000, weight_kg),
    height_cm = if_else(height_cm < 40, height_cm * 100, height_cm)
  )
```

---

### 3. Validity

To assess validity, you check whether values conform to what the field is
supposed to contain.

The regional bureau warns you about two columns where data capturers may have
made mistakes. First, sex comes in under several different codes, when it should
always be `F` or `M`. Second, woreda names come in with inconsistent
capitalisation and stray whitespace, when there should be exactly 57 names: 56
sampled woredas plus one "Unknown". You check both before fixing anything.

```{r validity-diagnose}
# Sex
count(cohort_clean, sex)

# Woredas
c(raw_names = n_distinct(cohort_clean$woreda_name),
  cleaned_names = n_distinct(str_to_title(str_squish(cohort_clean$woreda_name))))
```

Alongside fixing sex and woreda, you also convert age to a single unit, since
some rows record years and others months.

```{r validity-fix}
sex_lookup <- c("f" = "F", "female" = "F", "1" = "M",
                "m" = "M", "male" = "M", "2" = "F")

cohort_clean <- cohort_clean |>
  mutate(
    sex_std = unname(sex_lookup[str_to_lower(sex)]),
    woreda_std = str_to_title(str_squish(woreda_name)),
    age_months = if_else(age_unit == "months", age_value, age_value * 12)
  )
```

Note that `sex_std`, `woreda_std`, and `age_months` are new columns rather than
overwritten originals, so you can always get back to what was actually typed.

---

### 4. Consistency

To assess consistency, you check whether values that describe the same
real-world fact agree with each other. In this dataset, that means a child's
sex and date of birth should stay the same across every visit they made.

```{r consistency-diagnose}
cohort_clean |>
  group_by(person_id) |>
  summarise(n_sex = n_distinct(sex_std, na.rm = TRUE),
            n_dob = n_distinct(dob, na.rm = TRUE)) |>
  filter(n_sex > 1 | n_dob > 1) |>
  count(conflicting_sex = n_sex > 1, conflicting_dob = n_dob > 1)
```

Out of all the children, 30 have more than one recorded sex and 17 have more
than one recorded date of birth.

Note that this runs on `sex_std`, not `sex`. Without the standardisation from
the validity step above, `M`, `m` and `Male` would count as three different
values and almost every child would look inconsistent.

For each `person_id` with a conflict, you take the most common recorded value
across their visits and apply it everywhere. Note that this assumes the majority
entry is the correct one, which is a judgement call and should be documented.

```{r consistency-fix}
most_common <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return(NA)
  names(sort(table(x), decreasing = TRUE))[1]
}

fix_lookup <- cohort_clean |>
  group_by(person_id) |>
  summarise(sex_fix = most_common(sex_std),
            dob_fix = as.Date(most_common(as.character(dob))),
            .groups = "drop")

cohort_clean <- cohort_clean |>
  left_join(fix_lookup, by = "person_id") |>
  mutate(sex_std = sex_fix, dob = dob_fix) |>
  select(-sex_fix, -dob_fix)
```

---

### 5. Uniqueness

To assess uniqueness, you check whether the same event has been recorded more
than once. You start with the number of exact duplicate rows.

```{r uniqueness-diagnose}
sum(duplicated(cohort_clean))
```

You then check whether there are any people with repeated entries on the same
day.

```{r repeated-entries}
repeated_entries <- cohort_clean |>
  count(person_id, visit_date) |>
  filter(n > 1)

repeated_entries |> nrow()
```

You look at what an example of this repetition looks like for a single person.

```{r uniqueness-example}
cohort_clean |>
  filter(person_id == repeated_entries$person_id[1]) |>
  select(record_id, person_id, visit_date, weight_kg) |>
  arrange(visit_date)
```

Two fixes are needed. Identical rows are one submission sent twice, and
`distinct()` handles them cleanly. The rest are the same child on the same day
with a **new** `record_id` and a different weight. This is double data entry
where the attempts disagree. You keep the newer one, identified here by the
larger `record_id`, on the assumption that a value gets re-keyed because someone
noticed a problem with the first. As before, document the assumption.

```{r uniqueness-fix}
cohort_clean <- cohort_clean |>
  distinct() |>
  group_by(person_id, visit_date) |>
  slice_max(record_id, n = 1, with_ties = FALSE) |>
  ungroup()

nrow(cohort_clean)
```

---

### 6. Timeliness

To assess timeliness, you check whether data arrived, and was collected,
quickly enough to be useful.

The bureau tells you that they received data from most woredas in January, but
that Jimma Town only provided data in March. You calculate the reporting lag
(the gap in days between `visit_date` and `entry_date`) per woreda, and look at
the five woredas with the longest median lag.

```{r timeliness-diagnose}
cohort_clean |>
  mutate(lag_days = as.numeric(entry_date - visit_date)) |>
  group_by(woreda_std) |>
  summarise(median_lag = median(lag_days, na.rm = TRUE), .groups = "drop") |>
  arrange(desc(median_lag)) |>
  head(5)
```

Jimma Town has a median reporting lag of 97 days, about three months, while
every other woreda sits at 7 days, about a week, which aligns with what the
bureau told you. This means that most woredas report data soon after a visit,
but that Jimma Town may have a backlog or another issue preventing prompt
entry.

Since there is a reporting lag between a visit happening and it being entered,
you check the last month for which you have visit dates against the months
before it, in case the most recent month has not caught up yet.

```{r timeliness-month-diagnose}
cohort_clean |>
  mutate(month = floor_date(visit_date, "month")) |>
  count(month) |>
  arrange(month) |>
  tail(4)
```

The final month, December 2024, holds just 219 records, well under a
typical month's roughly 540. You assume it is a partial month and drop it.

```{r timeliness-fix}
cohort_clean <- cohort_clean |>
  mutate(month = floor_date(visit_date, "month")) |>
  filter(month < max(month))

nrow(cohort_clean)
```

---

### Document your pipeline

In this part you assessed each data dimension and fixed issues as they arose.
Another approach is to collect every fix into a single pipeline that cleans
everything in one place, which is easier to audit, rerun, and hand over to
someone else.

The same fixes, collected into a single pipeline, are given below:

```{r pipeline-single-block, eval=FALSE}
sex_lookup <- c("f" = "F", "female" = "F", "1" = "M",
                "m" = "M", "male" = "M", "2" = "F")

most_common <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return(NA)
  names(sort(table(x), decreasing = TRUE))[1]
}

cohort_clean <- read_csv("data/input/nutrition_screening.csv") |>
  # 1. Completeness: drop rows missing MCAR fields; leave muac_mm as is
  filter(!is.na(weight_kg), !is.na(height_cm)) |>
  # 2. Accuracy: repair confirmed unit errors in place
  mutate(
    age_unit = if_else(age_unit == "years" & age_value > 5, "months", age_unit),
    weight_kg = if_else(weight_kg > 30, weight_kg / 1000, weight_kg),
    height_cm = if_else(height_cm < 40, height_cm * 100, height_cm)
  ) |>
  # 3. Validity: standardise into new columns, keep the originals
  mutate(
    sex_std = unname(sex_lookup[str_to_lower(sex)]),
    woreda_std = str_to_title(str_squish(woreda_name)),
    age_months = if_else(age_unit == "months", age_value, age_value * 12)
  ) |>
  # 4. Consistency: resolve sex/dob to each person's majority value
  group_by(person_id) |>
  mutate(sex_std = most_common(sex_std),
         dob = as.Date(most_common(dob))) |>
  ungroup() |>
  # 5. Uniqueness: drop exact duplicates, keep the newest same-day entry
  distinct() |>
  group_by(person_id, visit_date) |>
  slice_max(record_id, n = 1, with_ties = FALSE) |>
  ungroup() |>
  # 6. Timeliness: drop the final, structurally incomplete month
  mutate(month = floor_date(visit_date, "month")) |>
  filter(month < max(month))
```

Whichever approach you take, write down what you did and what you assumed.

* **Completeness.** `weight_kg` and `height_cm` were confirmed MCAR and
  dropped; `muac_mm` was MAR but is not used in this analysis, so you did not
  fix it.
* **Accuracy.** Unit errors in weight, height and age repaired in place, no
  rows dropped.
  Assumption: every value past the threshold is a unit error rather than a
  genuine outlier, which rests on the bureau's confirmation.
* **Validity.** `sex_std`, `woreda_std` and `age_months` added as new columns,
  originals kept.
  Assumption: the codes `1`/`2` map to `M`/`F` as given, and title-casing
  plus trimming collapses every typo without merging two real woredas.
* **Consistency.** Sex and date of birth resolved by majority vote across a
  person's own visits.
  Assumption: the most frequent value is the correct one. When there is a tie,
  you take the earliest date, and `F` over `M` since it is alphabetically
  smaller.
* **Uniqueness.** Exact duplicates dropped, same-day conflicts resolved by
  `record_id`.
  Assumption: a larger ID on the same person and day means a corrected re-entry, not a second visit.
* **Timeliness.** The final incomplete month dropped; the slow woreda is a
  process finding, not a value to fix.
  Assumption: December's shortfall is a mid-month extract, not a real drop in visits.

---

## Part 2 - working with shapefiles

In this part you focus on working with shapefiles and joining your various
datasets to them. You will use two shapefiles. The first is supplied by the
regional bureau (`woredas.shp`) and contains the 56 woredas they sampled. The
second is `eth.shp`, which you downloaded yourself from the public national
database for country-wide context.

#### `woredas.shp` - 56 sampled woredas, UTM 37N (EPSG:32637)

| Field | Type | Description |
|---|---|---|
| `wor_pcode` | text | Woreda code |
| `wor_name` | text | Woreda name |
| `geometry` | MULTIPOLYGON | The woreda's boundary shape, in UTM 37N. |

```{r woredas-shape}
woredas <- st_read("data/input/woredas.shp", quiet = TRUE)
```

#### `eth.shp` - 1,148 national woredas, lon/lat (EPSG:4326)

Source: [HDX](https://data.humdata.org/dataset/cod-ab-eth). Woreda codes and
names match `woredas.shp`, and each woreda also carries the name and code of its
zone and region. Note that Oromia's zones have changed since the rainfall data was produced, with some
zones splitting. The shapefile has been edited to merge those zones back
together, so `zone_pcode` matches all 14 zones in the rainfall file.

| Field | Type | Description |
|---|---|---|
| `wor_pcode` | text | Woreda code |
| `wor_name` | text | Woreda name |
| `zone_pcode` | text | Zone code |
| `zone_name` | text | Zone name |
| `reg_pcode` | text | Region code |
| `reg_name` | text | Region name |
| `geometry` | MULTIPOLYGON | The woreda's boundary shape |

```{r eth-shape}
eth <- st_read("data/input/eth.shp", quiet = TRUE)
```

### 1. What is in a shapefile?

You start by printing the object and reading its header.

```{r what-is-sf}
woredas
```

The header is the useful part. **Geometry type** is MULTIPOLYGON, so some
woredas are in more than one piece. The **bounding box** numbers are in the
hundreds of thousands, so these are metres, not degrees. The **CRS** is UTM
zone 37N.

There are two things to watch for when you subset or summarise an sf object.

```{r sticky-geometry}
woredas |> select(wor_name)

eth |> st_drop_geometry() |> count(zone_name) |> head()
```

`select(wor_name)` returns two columns, not one, because geometry is **sticky**
and comes along unless you explicitly drop it. That is usually what you want on
a map and never what you want in a summary table. `woredas.shp` only has 56
rows, too few for that to cost you anything either way, so the second example
uses `eth`. 1,148 rows is where `st_drop_geometry()` first actually saves you
something, and where forgetting to drop it before a `group_by()` can turn
seconds into minutes.

---

### 2. Where is your data?

You add a flag to `eth` marking which woredas were sampled, so that you can see
your woredas in context.

```{r eth-enrich}
eth <- eth |>
  mutate(sampled = wor_pcode %in% woredas$wor_pcode)

eth |>
  st_drop_geometry() |>
  count(sampled)
```

This is an ordinary `mutate()` against a vector, not a join in the sf sense. It
matches on `wor_pcode`, a shared code, so it works even though `eth` is lon/lat
and `woredas` is UTM. CRS only matters once two geometries interact directly, as
Task 3 shows.

`eth` now carries everything you need: the national picture, the admin
hierarchy, and a flag for the 56 sampled woredas. You use `woredas` in Task 3
because it is the file with the wrong projection, but all other tasks use
`eth`, filtered to `sampled` when you mean just those 56.

Carrying the full country through every step costs more than it needs to.
Working from `woredas` and joining back to `eth` only at the end would be
faster, but keeping one object throughout is easier to follow.


You now plot some maps and look at the woredas your data comes from. You plot
the following five maps:

1. All woredas.

```{r plot-country, fig.height=5}
ggplot(eth) +
  geom_sf(fill = "gray90", colour = "black", linewidth = 0.1) +
  labs(title = "All woredas") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5))
```

2. Colour the woredas that are in Oromia.

```{r plot-region, fig.height=5}
ggplot(eth) +
  geom_sf(aes(fill = reg_name == "Oromia"), colour = "black", linewidth = 0.1) +
  scale_fill_manual(values = c("TRUE" = "#4A90D9", "FALSE" = "gray90")) +
  labs(title = "Woredas in Oromia", fill = "Oromia") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5))
```

3. Colour the woredas that were sampled.

```{r plot-sampled, fig.height=5}
ggplot(eth) +
  geom_sf(aes(fill = sampled), colour = "black", linewidth = 0.1) +
  scale_fill_manual(values = c("TRUE" = "#4A90D9", "FALSE" = "gray90")) +
  labs(title = "Sampled woredas", fill = "Sampled") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5))
```

4. Zoom into the Oromia region and plot the woredas that were sampled.

```{r plot-zoom, fig.height=5}
bb2 <- st_bbox(filter(eth, reg_name == "Oromia"))

ggplot(eth) +
  geom_sf(aes(fill = sampled), colour = "black", linewidth = 0.1) +
  scale_fill_manual(values = c("TRUE" = "#4A90D9", "FALSE" = "gray90")) +
  coord_sf(xlim = c(bb2$xmin, bb2$xmax), ylim = c(bb2$ymin, bb2$ymax)) +
  labs(title = "Sampled woredas, zoomed to Oromia", fill = "Sampled") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5))
```

5. Plot only the sampled woredas and nothing else.

```{r plot-sampled-only, fig.height=5}
ggplot(filter(eth, sampled)) +
  geom_sf(fill = "#4A90D9", colour = "gray90") +
  labs(title = "Sampled woredas only") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5))
```

**Note the difference between plot 4 and plot 5.** `coord_sf()` and `filter()`
answer different questions. `coord_sf()` restricts the view and keeps
neighbouring woredas visible at the edges. `filter()` deletes them, so the study
area floats in white space with no borders to locate it against. Zoom when you
want context; filter when you genuinely mean "these and only these".

**You need to choose a plot to show the sampled woredas in a report. Which
would you choose?** Either the third or the fourth plot. The fifth is not wrong,
but it removes the context that lets a reader judge coverage. Someone seeing
scattered fragments assumes that is the study area, rather than a sample
within a much larger region.

---

### 3. Coordinate reference systems

#### `health_posts.csv` - one row per health post

Locations of the health posts that captured data in the screening register
(`nutrition_screening.csv`).

| Field | Type | Description |
|---|---|---|
| `health_post` | text | Identifier. Joins to `nutrition_screening.csv`. |
| `hp_name` | text | Human-readable name. |
| `lon` | number | Longitude. A handful of rows have longitude and latitude swapped; see Part 2, Task 4. |
| `lat` | number | Latitude. A few rows are missing entirely; one sits exactly on a woreda boundary, by design. |

```{r hps}
posts <- read_csv("data/input/health_posts.csv")
```

You want to join the posts to one of your shapefiles so that you can see where
the health posts that collected your data are. You receive a csv with lon/lat
details, and you convert it into an `sf` object - geometry does not always have
to come from a shapefile.

```{r crs-check}
# 4326 = EPSG code for WGS84 (plain lon/lat)
posts_sf <- st_as_sf(posts, coords = c("lon", "lat"), crs = 4326)

posts_sf
```

Note that the geometry of `posts_sf` is a POINT.

Now you try to join `posts_sf` and `woredas`, which have different CRSs.

```{r crs-fail, error=TRUE}
st_join(posts_sf, woredas, join = st_within)
```

`sf` refuses: `st_crs(x) == st_crs(y) is not TRUE`. Metres and degrees are not
interchangeable, and a silent wrong answer would be worse than an error - every
point would land in the Gulf of Guinea, near (0, 0).

You transform the woredas into lon/lat so that both objects share a CRS.

```{r crs-fix}
woredas_ll <- st_transform(woredas, crs=4326)
st_crs(woredas_ll)$input
```

> **Note `st_transform()` vs `st_set_crs()`.** Transform recalculates every
> coordinate into the new system. Set just writes a different label onto
> unchanged numbers - appropriate only when the CRS was missing or wrong in the
> first place. Using `st_set_crs()` where `st_transform()` was meant moves
> the data thousands of kilometres and produces no error at all.

---

### 4. Point in polygon

You now join the posts to `woredas_ll` to find which polygon each health post
falls within. `st_join()` needs a predicate to decide what counts as a match.
You use `st_within`, which only counts a point as inside a polygon if it is
fully enclosed by it. The common alternative is `st_intersects`, which is more
generous: it counts anything touching the polygon at all, including its edge.

```{r spatial-join}
joined <- st_join(posts_sf, woredas_ll, join = st_within)

c(total_posts = nrow(joined),
  posts_matched = sum(!is.na(joined$wor_pcode)),
  posts_unmatched = sum(is.na(joined$wor_pcode)))
```

The order of the arguments matters: `st_join()` keeps the first object's rows and
geometry, so `posts_sf` has to come first for the result to have one row per
post. Swapping them gives one row per woreda instead, and
`st_within(woredas_ll, posts_sf)` would ask whether each polygon sits inside a
point.

Four posts were not matched. You plot the posts to see whether you can work out
why.

```{r plot-failures, fig.height=5}
ggplot() +
  geom_sf(data = eth, fill = "grey95", colour = "black", linewidth = 0.1) +
  geom_sf(data = woredas, fill = "blue", colour = "black") +
  geom_sf(data = st_jitter(joined, amount = 0.3), colour = "#D95F02", size = 1.5) +
  coord_sf(xlim = c(0, 46), ylim = c(0, 45)) +
  labs(title = "Health posts that landed nowhere", caption = sim_caption) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

Next you look at the longitude and latitude of the problematic posts.

```{r inspect-failures}
joined |>
  filter(is.na(wor_pcode)) |>
  # st_coordinates() pulls the X/Y values out of a geometry column into a
  # plain matrix - column 1 is X (lon), column 2 is Y (lat)
  mutate(lon = st_coordinates(geometry)[, 1],
         lat = st_coordinates(geometry)[, 2]) |>
  st_drop_geometry() |>
  select(health_post, lon, lat)

# Ethiopia's bounding box, for comparison
bb <- st_bbox(eth)
bb
```

Comparing these coordinates against Ethiopia's bounding box, it seems that the
lon and lat values have been entered the wrong way round, putting these posts in
the Mediterranean.

You flag the swapped rows, swap them back, and redo the join.

```{r fix-swapped}
posts_fixed <- posts |>
  mutate(
    swapped = lon < 20 & lat > 20,
    lon_fix = if_else(swapped, lat, lon),
    lat_fix = if_else(swapped, lon, lat)
  )

posts_sf2 <- posts_fixed |>
  st_as_sf(coords = c("lon_fix", "lat_fix"), crs = 4326)

joined2 <- st_join(posts_sf2, eth, join = st_within)

c(posts_unmatched = sum(is.na(joined2$wor_pcode)))
```

> **What if a post sits exactly on a woreda boundary?** `st_within` would not
> match it - a point on the edge is not strictly inside either polygon.
> `st_intersects` would match this point **twice**, since it touches the edge of
> both woredas. The geometry is ambiguous, and you would have to decide which
> behaviour you want.

---

### 5. Joining tables to a shapefile

#### `woreda_population.csv` - one row per woreda per year (2019-2024)

The number of people, and the number of under-5s, living in each woreda each
year.

| Field | Type | Description |
|---|---|---|
| `woreda_pcode` | text | Woreda code |
| `woreda_name` | text | Woreda name |
| `year` | number | Calendar year of the projection |
| `total_pop` | number | Total projected population |
| `pop_under2` | number | Projected population under 2 years |
| `pop_under5` | number | Projected population under 5 years |

```{r wor_pop}
pop <- read_csv("data/input/woreda_population.csv")
```

You can also join a table to a shapefile as normal, as long as there is a
column to join on.

```{r pop-join}
before <- nrow(woredas)
wor_pop <- woredas |> left_join(pop, by = c("wor_pcode" = "woreda_pcode"))

c(before = before,
  pop_years = n_distinct(pop$year),
  after = nrow(wor_pop))
```

The number of rows has grown, since each woreda now maps to each population year
(that is, the number of rows before multiplied by the number of population
years). The geometry is repeated across these rows, so exercise caution when
joining to shapefiles, as the output can grow quickly.

Now you try making the same join using woreda names.

```{r name-join}
by_name <- woredas |>
  left_join(pop, by = c("wor_name" = "woreda_name"))

sum(is.na(by_name$total_pop))

by_name |> filter(is.na(total_pop)) |> st_drop_geometry() |> select(wor_name)
```

Four woredas fail. The population file spells each with a trailing vowel
swapped for `e`: `Mulo` should have matched `Mule`, `Adama Tulu Jido
Kombolcha` should have matched `Adama Tulu Jido Kombolche`, `Bule Hora`
should have matched `Bule Hore`, and `Borecha` should have matched `Boreche`.
Joining by name is less reliable than using a code or coordinates.

---

### 6. Aggregating up

#### `zone_rainfall.csv` - one row per zone per month

Monthly rainfall for each zone, already summarised from satellite data.

| Field | Type | Description |
|---|---|---|
| `zone_pcode` | text | Zone unique identifier |
| `date` | date | First of the month that the rainfall applies to |
| `n_pixels` | number | Climate Hazards Group InfraRed Precipitation with Station data (CHIRPS) pixels averaged for this zone |
| `rain_mm` | number | Rainfall total for the month, in mm. |
| `rain_lta_mm` | number | Long-term average rainfall for that same calendar month, in mm |
| `rain_anom_pct` | number | `rain_mm` as a **percentage of `rain_lta_mm`**. 100 is normal, below 100 is drier than usual, above 100 is wetter |

```{r rainfall}
rainfall <- read_csv("data/input/zone_rainfall.csv")
```

Rainfall only exists at a zone level. You can either assume that each woreda
experiences the same amount of rain as the entire zone, or aggregate the
woredas to a zone level. Given that you need to answer policy questions related
to zones, you aggregate. You also keep track of which zones were sampled,
relative to the other zones.

```{r zones-grouped, fig.height=4}
zones <- eth |>
  group_by(zone_pcode, zone_name) |>
  summarise(sampled = ifelse(any(sampled), TRUE, FALSE),
            .groups = "drop")

nrow(zones)

ggplot(zones) +
  geom_sf(aes(fill = sampled), colour = "black", linewidth = 0.3, show.legend = FALSE) +
  labs(title = "Dissolved",
       subtitle = "14 zone shapes") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))
```

Note that `group_by() |> summarise()` on an sf object goes a step
further: it unions the geometries that share a group.

Now you merge the zones to the rainfall data and plot a single month.

```{r attach-rainfall, fig.height=5}
zone_rain <- zones |>
  left_join(rainfall, by = "zone_pcode")

# Rainfall for 2022-06-01
ggplot() +
  geom_sf(data = eth, fill = "gray90", colour = "black", linewidth = 0.2) +
  geom_sf(data = zone_rain |> filter(date == as.Date("2022-06-01")),
  aes(fill = rain_anom_pct), colour = "black", linewidth = 0.3) +
  scale_fill_viridis_c() +
  labs(title = "Rainfall, June 2022",
       subtitle = "Zone shapes are the dissolved sample, not the true extent",
       fill = "% of normal") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))
```

---

### 7. Tying it all together

You now need to tie the previous steps together and create a single object that
carries the screening records along with population and rainfall data.

You start with the screening records. You can either join them to `eth` at
woreda level on `woreda_name`, or you can go via the health post coordinates.
Joining on coordinates is more reliable than a name, and lets you reuse
`posts_sf2` from Part 2, Task 4 - the version with the swapped lon/lat rows
corrected and any post missing coordinates entirely dropped. To join using
coordinates, you first join to the health post object, then use `st_within()`
to place each record in a zone.

```{r attach-geography}
# cohort_clean was built in Part 1 and is still in memory; cohort is just
# the name this part works under, now that cleaning is done.
cohort <- cohort_clean

cohort_geo <- cohort |>
  left_join(posts_sf2 |> select(health_post), by = "health_post") |>
  st_as_sf()

cohort_geo_before <- nrow(cohort_geo)

# The first object's geometry is retained (the health post POINT), which
# you can drop. All you need is the zone code, which lets you re-attach the
# geometry whenever you want to plot.
combined_dataset <- cohort_geo |>
  st_join(eth |>
   select(wor_pcode, zone_pcode, zone_name, reg_pcode, reg_name),
   join = st_within) |>
  st_drop_geometry()

c(before = nrow(cohort_geo), after = nrow(combined_dataset))
```

`cohort` is a plain data frame, so joining it to the sf object `posts_sf2` keeps
the geometry column but not the `sf` class. `st_as_sf()` restores it. Because
`st_within()` is strict about boundaries, you check the row counts to confirm
that no points on a zone edge were dropped.

Finally, you attach population and rainfall to each visit.

```{r cohort-full}
rainfall <- rainfall |> rename(month = date)

# Remove the columns you don't need and remember to join on both the spatial
# identifier and time
combined_dataset <- combined_dataset |>
  mutate(year = year(visit_date)) |>
  left_join(pop |> select(-woreda_name), by = c("wor_pcode" = "woreda_pcode", "year")) |>
  left_join(rainfall |> select(-n_pixels), by = c("zone_pcode", "month"))

nrow(combined_dataset)
```

---

In this part you learnt how to work with shapefiles and join data to them.
Using what you have learned, you created `combined_dataset`, which you use in
the next part. It has one row per screening visit, with the rainfall for that
visit's zone that month and the population for its woreda that year attached to
each row.

**What this part has not covered:** small polygons a coarse grid misses
entirely, bounding boxes over-claiming on curved units, polygon-to-polygon
overlap and area weighting, and rasters. A tutorial covering these is in your
materials. When a spatial operation errors, try `st_is_valid()`,
`st_make_valid()`, `sf_use_s2(FALSE)` and `st_cast()` before anything else.

---

## Part 3 - building an outcome, and looking at it

### 1. From a weight to an outcome

To answer your policy question you need to know who is underweight. Raw weight
alone cannot tell you that. A 12 kg two-year-old and a 12 kg four-year-old are
not equally nourished, so weight only means something once it is compared
against what is expected for a child of that age and sex.

WHO's Child Growth Standards supply that reference, and the `anthro` package
turns each child's raw weight into a **weight-for-age z-score (WAZ)**. A WAZ
tells you how many standard deviations a child sits from the median child of
the same age and sex. According to WHO cutoffs, below -2 is underweight and
below -3 is severe.

`anthro_zscores()` is particular about its inputs. Sex must be coded `1`/`2`
rather than `"M"`/`"F"`, age must be in whole months, and a `measure` argument
tells it how height was taken. Children under 24 months are measured lying down
(`l`), older children standing up (`h`); the two give different numbers for the
same child, which the package adjusts for.

```{r zscores}
z <- anthro_zscores(
  sex = if_else(combined_dataset$sex_std == "M", 1, 2),
  age = combined_dataset$age_months,
  is_age_in_month = TRUE,
  weight = combined_dataset$weight_kg,
  lenhei = combined_dataset$height_cm,
  measure = if_else(combined_dataset$age_months < 24, "l", "h")
)

combined_dataset <- combined_dataset |>
  mutate(waz = z$zwei)

combined_dataset |>
  summarise(
    children = n(),
    no_zscore = sum(is.na(waz))
  )
```

---

### 2. Look at the outcome before you use it

Now look at a histogram of the outcome variable you have created, with dashed
lines at -2 and -3, the WHO thresholds for underweight and severely underweight
respectively.

```{r hist, fig.height=4}
combined_dataset |>
  ggplot(aes(waz)) +
  geom_histogram(bins = 60, fill = "#4A90D9") +
  geom_vline(xintercept = c(-3, -2), linetype = "dashed", colour = "#D95F02") +
  labs(title = "Weight-for-age z-scores",
       subtitle = "Dashed lines: underweight (-2) and severe (-3)",
       x = "WAZ", y = "Children", caption = sim_caption) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        panel.border = element_blank(),
        axis.line = element_line(colour = "black"))
```

The curve is roughly normal, sitting to the left of zero, with a tail
stretching past -2. That tail is the children who are underweight. Next you
create a boxplot that splits the same distribution by zone.

```{r box, fig.height=4}
combined_dataset |>
  ggplot(aes(x = fct_reorder(zone_name, waz), y = waz)) +
  geom_boxplot(fill = "#4A90D9", alpha = 0.6, outlier.alpha = 0.2) +
  geom_hline(yintercept = c(-3, -2), linetype = "dashed", colour = "#D95F02") +
  coord_flip() +
  labs(title = "Weight-for-age by zone", x = NULL, y = "WAZ",
       caption = sim_caption) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        panel.border = element_blank(),
        axis.line = element_line(colour = "black"))
```

Every zone has the same left tail, so the skew is not one bad zone dragging the
rest down. Most boxes overlap heavily, with two exceptions: Ilu Aba Bora sits
clearly below the others, and Arsi sits above.

---

### 3. Mapping the children underweight

While the z-score is useful, you are more interested in the number of children
who are underweight and who are severely underweight.

```{r underweight-flags}
# Moderate and severe are exclusive bands; total_underweight is everyone
# below -2, which is the WHO "underweight" definition.
combined_dataset <- combined_dataset |>
  mutate(
    underweight = -3 <= waz & waz < -2,
    severe = waz < -3,
    total_underweight = underweight | severe
  )

combined_dataset |>
  summarise(
    pct_underweight = round(100 * mean(underweight), 1),
    pct_severe = round(100 * mean(severe), 1),
    pct_total_underweight = round(100 * mean(total_underweight), 1)
  )
```

Now you plot your data. You consider three views: the number of children
screened, the number underweight, and the proportion of screened children who
are underweight. First you build the zone-level summary those maps need.

```{r by-zone}
# pop_under5 is a per-woreda figure, repeated on every child's visit row in
# combined_dataset - summing it directly would add that same number in once
# per visit instead of once per woreda, so take distinct woreda values first.
zone_pop <- combined_dataset |>
  distinct(wor_pcode, zone_pcode, pop_under5) |>
  group_by(zone_pcode) |>
  summarise(pop_under5 = sum(pop_under5, na.rm = TRUE), .groups = "drop")

zone_summary <- combined_dataset |>
  group_by(zone_pcode) |>
  summarise(
    n_screened = n(),
    n_underweight = sum(underweight),
    n_severe = sum(severe),
    n_total_underweight = sum(total_underweight),
    prop_screened_underweight = n_total_underweight / n_screened,
    .groups = "drop"
  ) |>
  left_join(zone_pop, by = "zone_pcode")

zone_data <- zones |> left_join(zone_summary, by = "zone_pcode")
```

You define a helper function to speed up the plotting, then use it to create
three maps.

```{r three-maps, fig.width=7, fig.height=15, warning=FALSE}
p <- function(var, title, fill_title, labels = scales::label_number()) {
  bb <- st_bbox(filter(zone_data, sampled == TRUE))

  ggplot() +
    geom_sf(data = zone_data, aes(fill = .data[[var]]), colour = "black", linewidth = 0.2) +
    geom_sf_label(data = filter(zone_data, sampled), aes(label = zone_name), size = 2,
                  linewidth = 0, fill = scales::alpha("gray90", 0.7)) +
    # to zoom in, uncomment this line
    # coord_sf(xlim = bb[c("xmin", "xmax")], ylim = bb[c("ymin", "ymax")]) +
    scale_fill_viridis_c(direction = -1, na.value = "grey95", labels = labels) +
    labs(title = title, fill = fill_title, caption = sim_caption) +
    theme_void() +
    theme(plot.title = element_text(hjust = 0.5))
}

p("n_total_underweight", "a) Underweight children", "Children") +
p("n_screened", "b) Children screened", "Children") +
p("prop_screened_underweight", "c) Proportion underweight", "Proportion",
  labels = scales::label_percent()) +
  plot_layout(ncol = 1)
```

**Compare Ilu Aba Bora and Arsi.** Map (a) shows the number of underweight
children and, read alone, it suggests that these are the worst-off zones in
Oromia. But map (b) shows that the number of children screened is markedly
different: Arsi screened the most of any zone, Ilu Aba Bora among the fewest.
Map (c) divides one by the other. Ilu Aba Bora has the highest prevalence; Arsi
falls to the bottom. Its count was high only because its denominator was.

Counts and prevalence are not the same thing, and confusing them is a common
mistake. You need a denominator, especially when comparing across spatial units
or time.

Colour on a map is hard to rank precisely. The same children as bars, sorted and
split by severity, make the point more clearly:

```{r severity-bars, fig.height=9}
zone_severity <- zone_data |>
  st_drop_geometry() |>
  filter(sampled) |>
  select(zone_name, n_screened, Underweight = n_underweight, Severe = n_severe) |>
  pivot_longer(c(Underweight, Severe), names_to = "severity", values_to = "n") |>
  mutate(
    severity = factor(severity, levels = c("Underweight", "Severe")),
    prop_screened = n / n_screened
  )

q <- function(value_var, title, y_lab, labels = scales::label_number()) {
  ggplot(zone_severity, aes(x = fct_reorder(zone_name, .data[[value_var]], sum),
                            y = .data[[value_var]], fill = severity)) +
    geom_col() +
    coord_flip() +
    scale_y_continuous(labels = labels) +
    scale_fill_manual(values = c(Underweight = "#4A90D9", Severe = "#D95F02")) +
    labs(title = title, x = NULL, y = y_lab, fill = NULL, caption = sim_caption) +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5),
          panel.border = element_blank(),
          axis.line = element_line(colour = "black"))
}

q("n", "Underweight children",
  "Children") +
q("prop_screened", "Proportion of children screened",
  "Proportion", labels = scales::label_percent()) +
  plot_layout(ncol = 1)
```

---

### 4. Coverage

Map (c) ranks zones by the proportion underweight among children screened.
But that proportion only describes the children who came, not the children who
live there. To take this into account you need to investigate coverage. You add
coverage to `zone_data` and map it alongside the under-5 population.

```{r coverage, fig.width=7, fig.height=10}
zone_data <- zone_data |>
  mutate(coverage = n_screened / pop_under5)

p("pop_under5", "d) Under-5 population",
  "Children under-5") +
p("coverage", "e) Coverage (screened ÷ under-5 population)",
  "Screened", labels = scales::label_percent(accuracy = 0.1)) +
  plot_layout(ncol = 1)
```

Coverage is extremely poor across all zones. Arsi screened just 0.90% of its
under-5 population, the highest in the sample; West Hararge screened 0.31%, the
lowest, despite having the largest under-5 population in the region.

Ranking the zones by the proportion underweight, alongside their coverage,
gives:

```{r final-ranking}
zone_data |>
  st_drop_geometry() |>
  filter(sampled) |>
  arrange(desc(prop_screened_underweight)) |>
  transmute(
    zone_name,
    n_screened,
    n_total_underweight,
    pct_underweight = round(100 * n_total_underweight / n_screened, 1),
    pct_coverage = round(100 * coverage, 2)
  )
```

**So, which zones in Oromia have the highest malnutrition?** After adjusting for
the number of children screened, Ilu Aba Bora, East Hararge, Guji and Borena
have the highest proportion of children underweight. In contrast, Arsi, North
Shewa (OR), Jimma, and West Wellega have the lowest.

However, there are limitations that need to be acknowledged when interpreting
these maps. Only a small proportion of children under 5 in the region are
reached, with coverage under 1% everywhere, so none of these figures estimates
prevalence in the population. You have also used a sample of woredas to
represent each entire zone. Screening may not be random either; it is possible
that caregivers bring in the children who already look sick, which pushes every
zone's proportion above its true prevalence by an unknown amount. To understand
this better you would need to ask the bureau for more details about how this
sample was collected, and possibly improve the sampling process going forward
if it is not possible to increase coverage.

---

### 5. Over time

Now you need to address question 2. As a reminder, `rain_anom_pct` is that
month's rainfall as a percentage of the long-term average for that same month
in that same zone: 100 is normal, below 100 is drier than usual, above 100 is
wetter. This is the same idea as the proportions above. Absolute rainfall is not
comparable across seasons or zones, so it needs a denominator too, and here the
denominator is what that zone normally gets at that time of year.

Start by looking at the proportion underweight and `rain_anom_pct` overall.

```{r time-plot, fig.height=5}
monthly <- combined_dataset |>
  group_by(month) |>
  summarise(n_screened = n(),
            n_total_underweight = sum(total_underweight),
            prop = n_total_underweight / n_screened,
            rain_pct = mean(rain_anom_pct, na.rm = TRUE),
            .groups = "drop")

p_rain <- ggplot(monthly, aes(month, rain_pct)) +
  geom_line(colour = "#4A90D9", linewidth = 0.9) +
  labs(title = "Nutrition follows the rains, with a delay",
       x = NULL, y = "Rainfall, % of normal") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        panel.border = element_blank(),
        axis.line = element_line(colour = "black"))

p_prop <- ggplot(monthly, aes(month, prop)) +
  geom_line(colour = "#D95F02", linewidth = 0.9) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = NULL, y = "Proportion underweight", caption = sim_caption) +
  theme_bw() +
  theme(panel.border = element_blank(),
        axis.line = element_line(colour = "black"))

p_rain / p_prop
```

Rainfall (top) dips through 2021 and 2022, the sharpest part of the Horn of
Africa drought (2020-2023) as it shows up in this data, while the proportion
of children underweight (bottom) rises afterwards rather than at the same time. One explanation for the lag is the chain from poor
rains to crop and livestock losses, then to household food insecurity, and
finally to malnourishment. It is important to be clear, though, that nothing
here establishes that chain. Two series moving in sequence is consistent with
it, but it may also support many other explanations.

**How long is that delay?** Rather than guess, you check every lag from 0 to 9
months and see which one actually correlates with the proportion underweight.

```{r lag-scan, fig.height=4}
lag_scan <- data.frame(lag = 0:9) |>
  mutate(correlation = sapply(lag, function(k) {
    cor(monthly$prop, lag(monthly$rain_pct, k), use = "complete.obs")
  }))

ggplot(lag_scan, aes(lag, correlation)) +
  geom_col(fill = "#4A90D9") +
  geom_hline(yintercept = 0, colour = "black") +
  scale_x_continuous(breaks = 0:9) +
  labs(title = "How many months back does rainfall predict underweight?",
       x = "Lag (months)", y = "Correlation with proportion underweight",
       caption = sim_caption) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        panel.border = element_blank(),
        axis.line = element_line(colour = "black"))
```

Rain in the same month, or the month before, tells you almost nothing -
correlation is close to zero through lag 2. It jumps sharply at 3 months, stays
strong through 6, then fades back towards zero by 7. From here you use the
average of lags 3-6, `rain_lag`, rather than picking a single month out of that
window.

Based on the above, you plot the average rainfall at a lag of 3-6 months, and
overlay the two series so that they move together. The relationship should be
clearer now.

```{r lag-overlay, fig.height=4.5}
monthly_shifted <- monthly |>
  mutate(rain_lag_pct = (lag(rain_pct, 3) + lag(rain_pct, 4) +
                          lag(rain_pct, 5) + lag(rain_pct, 6)) / 4)

scale_factor <- max(monthly_shifted$rain_lag_pct, na.rm = TRUE) /
  max(monthly_shifted$prop, na.rm = TRUE)

ggplot(monthly_shifted, aes(month)) +
  geom_line(aes(y = rain_lag_pct), colour = "#4A90D9", linewidth = 0.9, na.rm = TRUE) +
  geom_line(aes(y = prop * scale_factor), colour = "#D95F02", linewidth = 0.9) +
  scale_y_continuous(
    name = "Rainfall 3-6 months earlier (% of normal)",
    sec.axis = sec_axis(~ . / scale_factor, name = "Proportion underweight",
                         labels = scales::percent)
  ) +
  labs(title = "Shifted by the lag, the two lines move together",
       subtitle = "Blue: rainfall 3-6 months earlier (left axis). Orange: proportion underweight (right axis).",
       x = NULL, caption = sim_caption) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        axis.title.y.left = element_text(colour = "#4A90D9"),
        axis.title.y.right = element_text(colour = "#D95F02"),
        panel.border = element_blank(),
        axis.line.x.bottom = element_line(colour = "black"),
        axis.line.y.left = element_line(colour = "black"),
        axis.line.y.right = element_line(colour = "black"))
```

---

### 6. The relationship

Now that you know there is a relationship overall, look at it at a zone level.
You start by creating a `zone_month` object, which is similar to `monthly` but
also groups by `zone_pcode` and `zone_name`.

```{r zone-month}
zone_month <- combined_dataset |>
  group_by(zone_pcode, zone_name, month) |>
  summarise(n_screened = n(),
            n_total_underweight = sum(total_underweight),
            prop = n_total_underweight / n_screened,
            rain_anom_pct = mean(rain_anom_pct),
            .groups = "drop")

zone_month <- zone_month |>
  arrange(zone_pcode, month) |>
  group_by(zone_pcode) |>
  mutate(rain_lag = (lag(rain_anom_pct, 3) + lag(rain_anom_pct, 4) +
                     lag(rain_anom_pct, 5) + lag(rain_anom_pct, 6)) / 4) |>
  ungroup() |>
  filter(!is.na(rain_lag)) # lagging costs you the first six months of data
```

You start by looking at a scatter of lagged rainfall against the proportion
underweight.

```{r scatter, fig.height=4}
zone_month |>
  filter(n_screened >= 25) |>
  ggplot(aes(rain_lag, prop)) +
  geom_point(aes(), alpha = 0.25, colour = "#4A90D9") +
  geom_vline(xintercept = 100, linetype = "dotted", colour = "black") +
  geom_smooth(aes(weight = n_screened), method = "glm",
              method.args = list(family = quasibinomial()),
              se = FALSE, colour = "#D95F02") +
  labs(title = "Underweight against the previous season's rainfall",
       subtitle = "Each point is a zone-month; below 100% is drier than normal",
       x = "Rainfall 3-6 months earlier (% of normal)",
       y = "Proportion underweight", caption = sim_caption) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        panel.border = element_blank(),
        axis.line = element_line(colour = "black"))
```

There is a negative relationship between rainfall and the proportion of children
underweight; that is, as rainfall decreases, the proportion underweight
increases. The effect is more pronounced below 100%, where the curve steepens,
while above about 150% it flattens out and extra rain makes little difference.
The scatter around the curve is wide, so this describes an average pattern
rather than something that would predict any single zone-month.

**Is that relationship stable, or is it one drought year doing all the work?**
To answer this, look at the correlation within each year.

```{r yearly-corr, fig.height=4}
yearly_corr <- zone_month |>
  mutate(year = year(month)) |>
  group_by(year) |>
  # Neither prop nor rain_lag has NAs left at this point, so this cor() is
  # safe as-is. If they did, cor()'s use = "complete.obs" argument would drop
  # incomplete pairs instead of returning NA.
  summarise(correlation = cor(prop, rain_lag),
            n_obs = sum(!is.na(prop) & !is.na(rain_lag)),
            .groups = "drop")

ggplot(yearly_corr, aes(factor(year), correlation)) +
  geom_col(fill = "#4A90D9") +
  geom_hline(yintercept = 0, colour = "black") +
  geom_text(aes(label = paste0("n=", n_obs),
                y = pmin(correlation, 0) - 0.03),
            size = 3, colour = "grey30") +
  labs(title = "The rain-underweight correlation, year by year",
       subtitle = "n = zone-month observations feeding each year's correlation",
       x = NULL, y = "Correlation (prop underweight vs. 3-6 month rain lag)",
       caption = sim_caption) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        panel.border = element_blank(),
        axis.line = element_line(colour = "black"))
```

The correlation is negative in every year and never weaker than about -0.6, so
no single year is driving the result. 2019 is the strongest year at around
-0.82, but note that it has fewer datapoints, since lagging costs you the first
six months of the series. That leaves only part of the year, and because both rainfall and the
proportion underweight vary seasonally, a partial year is not directly
comparable with a full one.

**Is it the same in every zone, or concentrated in a few?**

```{r zone-corr-map, fig.height=5, warning=FALSE}
zone_corr <- zone_month |>
  group_by(zone_pcode) |>
  # Same as above: safe without use = "complete.obs" here, since prop and
  # rain_lag are already complete, but that argument is what you'd reach for
  # if either one had NAs.
  summarise(n_months = n(),
            correlation = cor(prop, rain_lag),
            .groups = "drop")

zone_corr_map <- zones |>
  left_join(zone_corr, by = "zone_pcode") |>
  filter(!is.na(correlation))

# Zoom to the zones that actually carry a correlation. `bb` further up is
# Ethiopia's national bounding box and would zoom out to the whole country.
bb_corr <- st_bbox(zone_corr_map)

ggplot() +
  geom_sf(data = zones, fill = "grey95", colour = "black", linewidth = 0.1) +
  geom_sf(data = zone_corr_map, aes(fill = correlation), colour = "black", linewidth = 0.2) +
  geom_sf_label(data = zone_corr_map, aes(label = zone_name), size = 2,
                linewidth = 0, fill = scales::alpha("gray90", 0.7)) +
  scale_fill_viridis_c(limits = c(-0.6, -1)) +
  coord_sf(xlim = bb_corr[c("xmin", "xmax")], ylim = bb_corr[c("ymin", "ymax")]) +
  labs(title = "How rain-sensitive is each zone?",
       subtitle = "Pearson correlation between lagged rainfall and proportion underweight, pooled across all six years",
       fill = "Pearson correlation", caption = sim_caption) +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))
```

Note that a correlation describes how two series move together. It says nothing
about how many children are underweight in a zone, so these values cannot be
used to rank zones as better or worse off. The two are separate, and reading them together is
what makes the zone comparisons informative.

Every zone is negative, but not equally so, ranging from about -0.63 in North
Shewa (OR) to -0.91 in Borena. Borena, Guji and East Hararge rank high on both: a high
proportion of children under 5 are underweight, and their series track rainfall
closely. These zones are therefore both worse off and more exposed to variation
in rainfall. Ilu Aba Bora is the contrast. It has the highest underweight proportion in the region, yet its
series moves less closely with rainfall, so other factors are likely contributing.

**So, is rainfall associated with child malnutrition in Oromia?** Yes, there is
a negative relationship between rainfall and the proportion of children under 5
who are underweight. However, that relationship does not hold in the same way
everywhere: Ilu Aba Bora's problem looks chronic rather than rain-driven, and
Arsi's rain-sensitivity does not show up in its baseline numbers at all, so
rainfall alone does not explain either zone. A further limitation of this
correlation view is that each zone's number is built from a small sample (57-65
zone-months).

---

## Before you leave

With that, you have enough to send the Regional Health Bureau a preliminary
reply. They asked two questions, and you now have two answers.

1. **Which zones in Oromia have the highest malnutrition?**
Ilu Aba Bora, East Hararge, Guji and Borena - all at or above 20% of screened
children underweight, roughly double the lowest zone, West Wellega, at under 11%
(Part 3.4). Three of those four worst zones
are also among the most rain-sensitive, which is the shape a genuine
drought-vulnerability story predicts.

2. **Is rainfall associated with child malnutrition in Oromia?**
Yes, as rainfall decreases the proportion of underweight children increases.
Borena, Guji and East Hararge stand out as most at risk - they combine a high
proportion of underweight children with series that track rainfall closely.

Neither answer should be reported without its limitations. Coverage is under 1%
in every zone, the woredas are a sample standing in for whole zones, and the
children screened are not a random sample of the children who live there.
Reporting these caveats alongside the findings is what allows readers to
interpret them correctly and use them responsibly.

There is also a limitation in the approach you have taken. You have examined
the outcome over space and time, and asked whether a single variable (rainfall)
is associated with it while considering nothing else. The analysis shows only
that the two move together; it does not quantify how much underweight rises for
a given shortfall in rainfall. In practice the drivers of public health
outcomes are multifactorial, and a variable can appear important simply because
it is correlated with the one that actually matters. Separating these effects
requires a model that considers them jointly, which is what the next session
covers. Once you have worked through it, return to this dataset and see what
changes.
