---
title: "EDAM training - cleaning and exporing data (exercises)"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
    number_sections: false
    theme: flatly
    highlight: tango
    code_folding: show
    df_print: paged
    fig_width: 8
    fig_height: 5
---

> ⚠️ **This is simulated data, for teaching purposes only.**
> All screening, population, and health post data is simulated.
> Nothing here describes a real person, health facility, programme or nutrition outcome.
> Only the rainfall data is real, sourced from [OCHA HDX](https://data.humdata.org/dataset/eth-rainfall-subnational).

## How to use this notebook

This is the exercise version of session 1. A few things to know before you start:

- Wherever you see `___` inside a code chunk, that's something for you to fill in. A `# TODO:`
  comment above it explains what's needed.
- A `___` is not valid R on its own, so a chunk with a blank in it **will error** if you try to
  run it as-is. That's expected. Work through this chunk by chunk (Ctrl/Cmd+Enter to run just the
  current chunk) as you fill things in, rather than trying to Knit the whole document until you've
  completed everything.
- Wherever you see a **bolded question**, work out your own answer from the code and output above
  it before moving on. The answers aren't in this file — check your reasoning against
  `session_1_SOLUTIONS.Rmd` (or the published walkthrough on the course site) when you're done, or
  if you get stuck.

## Overview

In this tutorial, imagine you work in an analysis team in a nutrition
office in the Ethiopian Ministry of Health. You have just received a
request from the Oromia Regional Health Bureau. Rainfall has been poor
across the region for the past two seasons, and they want to know whether
this is showing up in child nutrition. Based on their request, you
determine that there are two main questions to answer:

1. **Which zones in Oromia have the highest malnutrition?**
2. **Is rainfall associated with child malnutrition in Oromia?**

To answer these you have growth-monitoring records for children under 5 from
health posts across 56 woredas (districts) in 14 zones, and rainfall estimates
for the same zones and months. There is no budget or time for a survey.

You will work towards an answer in three parts, each building on the
output of the last:

1. **Checking data quality.** You apply six data quality dimensions to the raw
   register: accuracy, completeness, uniqueness, consistency, timeliness, and
   validity. This gives you one documented cleaning pipeline.
2. **Working with shapefiles.** This part focuses on shapefiles and, more
   specifically, on how you get the cleaned register onto a map. It covers
   coordinate reference systems, spatial joins, and the difference between
   dissolving geometry and aggregating data.
3. **Creating an outcome variable and exploring it.** You determine how many
   children are underweight, and convert these to proportions so you can
   compare across zones. You then use exploratory analysis to inspect these
   proportions and answer both questions above.

By the end you will have a clean, geographically located, outcome-scored
dataset, and an answer to both questions along with its limitations.

## Setup

> **Note:** This notebook will not knit until every `___` blank above has been
> filled in - a chunk containing a blank is not valid R and will error the
> moment it is evaluated.

```{r, echo=FALSE}
knitr::opts_chunk$set(fig.align = "center")
```

```{r setup, message=FALSE, warning=FALSE}
# Setup chunk
# Libraries used in this tutorial
library(anthro)
library(dplyr)
library(forcats)
library(ggplot2)
library(lubridate)
library(patchwork)
library(readr)
library(sf)
library(stringr)
library(tidyr)

# Caption for plots
sim_caption <- "*Simulated data - for teaching purposes only"
```

## Part 1 - Check your ingredients

Throughout this tutorial you will use four data tables and two shapefiles.
You read in each file as it is required, and a brief explanation of the
dataset and its codebook comes before each one.

In this part, you will focus on nutrition screening as an illustrative example, but in
practice you should check all your files.

#### `nutrition_screening.csv` - one row per child per visit

The raw growth-monitoring register. This captures every visit a child made to a
health post, typed in by hand.

| Field | Type | Description |
|---|---|---|
| `record_id` | text | Unique visit identifier |
| `person_id` | text | Unique person identifier (identifies the person across visits) |
| `woreda_name` | text | Woreda name as typed at the health post |
| `health_post` | text | Unique identifier of the health post the child attended |
| `visit_date` | date | Date of the screening visit |
| `entry_date` | date | Date the record was entered into the system. Compare to `visit_date` for reporting lag |
| `sex` | text | The sex of the child |
| `dob` | date | Date of birth of the child |
| `age_value` | number | Age as recorded, in the unit given by `age_unit` |
| `age_unit` | text | The age unit (`"months"`/`"years"`) |
| `weight_kg` | number | Child's weight in kilograms (kg) |
| `height_cm` | number | Child's height/length in centimetres (cm) |
| `muac_mm` | number | Mid-upper arm circumference (MUAC), in millimetres (mm) |
```{r screening}
screening <- read_csv("data/nutrition_screening.csv")
```

### What have you got?

You start by looking at the first few rows, and then at a summary of every
column.

```{r shape}
head(screening)
summary(screening)
```

**Question:** What data quality issues can you already spot in the output above?

Every check below works on `cohort_clean` - a single object you build up as you
go, one dimension at a time. Where a dimension has a known, explainable fix,
you apply it immediately and `cohort_clean` carries the fix into the next
check. At the end you collect the same set of decisions into one documented
pipeline, so you can see both ways of getting there.

```{r cohort-init}
cohort_clean <- screening
```

---

### 1. Completeness

To assess completeness, you check what percentage of values are missing per
column and look for patterns.

```{r completeness-a}
cohort_clean |>
  summarise(across(everything(), ~ round(mean(is.na(.)) * 100, 1))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "pct_missing") |>
  arrange(desc(pct_missing))
```

You raise `weight_kg`, `height_cm`, `sex` and `muac_mm` with the regional bureau.
They come back having already looked into three of the four: `weight_kg` and
`height_cm` are **missing completely at random (MCAR)**, meaning the missingness
is unrelated to the data. There is nothing to explain, and nothing to fix beyond
dropping those rows. However, `sex` is missing at the visit level rather than the
child level - a child's other visits usually carry a usable value. These rows
are recovered when we consider consistency.

```{r completeness-drop}
# TODO: weight_kg and height_cm were confirmed MCAR by the bureau - drop rows
# where either one is missing.
cohort_clean <- cohort_clean |>
  filter(!is.na(___), ___)

nrow(cohort_clean)
```

They are less sure about `muac_mm`, though. Their guess is **missing at random
(MAR)**, meaning it is missing for a reason tied to something you can see in
the data. They suggest looking at the health post, since that is the level at
which equipment and staffing actually vary:

```{r completeness-b}
cohort_clean |>
  group_by(health_post) |>
  summarise(muac_missing = round(100 * mean(is.na(muac_mm))), .groups = "drop") |>
  filter(muac_missing > 30) |>
  arrange(desc(muac_missing))
```

**Question:** Does the missingness above look evenly spread across health posts, or does it
cluster? What does that suggest about the mechanism, and does it match the bureau's MAR guess?

`muac_mm` is not used anywhere else in this analysis, so you do not repair this
missingness.

**Since you take this to be MAR, how would you address the missingness?**

**What would you need to see instead for `muac_mm` to be MNAR? Give a concrete
example.**

---

### 2. Accuracy

To assess accuracy, you check whether recorded values are physically possible,
and wherever you know the actual mechanism behind the error, you repair it
rather than remove it.

You contact the regional office that provided this data for guidance. They
confirm three known issues: some weights were mistakenly recorded in grams
rather than kilograms, some heights in metres rather than centimetres, and some
ages in months have been recorded as years. You count how many rows each issue
affects.

```{r accuracy-diagnose}
cohort_clean |>
  summarise(
    weight_bad = sum(weight_kg > 30, na.rm = TRUE),
    height_bad = sum(height_cm < 40, na.rm = TRUE),
    age_bad = sum(age_unit == "years" & age_value > 5, na.rm = TRUE)
  )
```

You then look at a few of the affected rows.

```{r accuracy-values}
cohort_clean |>
  filter(weight_kg > 30 | height_cm < 40 | (age_unit == "years" & age_value > 5)) |>
  select(age_value, age_unit, weight_kg, height_cm) |>
  head(8)
```

Given that you know the cause, you can fix the problematic rows. If you did not
know the cause, you would need to decide whether to drop the rows (assuming
there is no pattern to which ones are problematic) or contact the bureau to
figure out the cause.

```{r accuracy-fix}
# TODO: repeat the three conditions from accuracy-diagnose above to fix each
# issue in place - weight_kg over 30 is really grams (divide by 1000);
# height_cm under 40 is really metres (multiply by 100); age_unit == "years"
# together with age_value > 5 is really months.
cohort_clean <- cohort_clean |>
  mutate(
    age_unit = if_else(___, "months", age_unit),
    weight_kg = if_else(___, weight_kg / 1000, weight_kg),
    height_cm = if_else(___, ___, ___)
  )
```

---

### 3. Validity

To assess validity, you check whether values conform to what the field is
supposed to contain.

The regional bureau warns you about two columns where data capturers may have
made mistakes. First, sex comes in under several different codes, when it should
always be `F` or `M`. Second, woreda names come in with inconsistent
capitalisation and stray whitespace, when there should be exactly 57 names: 56
sampled woredas plus one "Unknown". You check both before fixing anything.

```{r validity-diagnose}
# Sex
count(cohort_clean, sex)

# Woredas
c(raw_names = n_distinct(cohort_clean$woreda_name),
  cleaned_names = n_distinct(str_to_title(str_squish(cohort_clean$woreda_name))))
```

Alongside fixing sex and woreda, you also convert age to a single unit, since
some rows record years and others months.

```{r validity-fix}
# TODO: sex comes in as f/female/2 (female) and m/male/1 (male). Map each
# code onto "F" or "M".
sex_lookup <- c(___ = ___, "female" = ___, ___ = "M",
                ___ = ___, "male" = ___, ___ = "F")

cohort_clean <- cohort_clean |>
  mutate(
    sex_std = unname(sex_lookup[str_to_lower(sex)]),
    # TODO: woreda_std should be woreda_name in title case, with extra/stray
    # whitespace squished out
    woreda_std = ___,
    # TODO: age_months should equal age_value when age_unit is "months", and
    # age_value * 12 when age_unit is "years"
    age_months = ___
  )
```

Note that `sex_std`, `woreda_std`, and `age_months` are new columns rather than
overwritten originals, so you can always get back to what was actually typed.

---

### 4. Consistency

To assess consistency, you check whether values that describe the same
real-world fact agree with each other. In this dataset, that means a child's
sex and date of birth should stay the same across every visit they made.

```{r consistency-diagnose}
cohort_clean |>
  group_by(person_id) |>
  summarise(n_sex = n_distinct(sex_std, na.rm = TRUE),
            n_dob = n_distinct(dob, na.rm = TRUE)) |>
  filter(n_sex > 1 | n_dob > 1) |>
  count(conflicting_sex = n_sex > 1, conflicting_dob = n_dob > 1)
```

Note that this runs on `sex_std`, not `sex`. Without the standardisation from
the validity step above, `M`, `m` and `Male` would count as three different
values and almost every child would look inconsistent.

For each `person_id` with a conflict, you take the most common recorded value
across their visits and apply it everywhere. Note that this assumes the majority
entry is the correct one, which is a judgement call and should be documented.

```{r consistency-fix}
most_common <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return(NA)
  names(sort(table(x), decreasing = TRUE))[1]
}

# TODO: for each person_id, find their majority sex_std and dob (as a Date)
# across all of their visits, using most_common() above.
fix_lookup <- cohort_clean |>
  group_by(person_id) |>
  summarise(sex_fix = ___,
            dob_fix = ___,
            .groups = "drop")

# TODO: join fix_lookup back on, and overwrite sex_std/dob with the majority
# values you just computed.
cohort_clean <- cohort_clean |>
  left_join(fix_lookup, by = "person_id") |>
  mutate(sex_std = ___, dob = ___) |>
  select(-sex_fix, -dob_fix)
```

---

### 5. Uniqueness

To assess uniqueness, you check whether the same event has been recorded more
than once. You start with the number of exact duplicate rows.

```{r uniqueness-diagnose}
sum(duplicated(cohort_clean))
```

You then check whether there are any people with repeated entries on the same
day.

```{r repeated-entries}
repeated_entries <- cohort_clean |>
  count(person_id, visit_date) |>
  filter(n > 1)

repeated_entries |> nrow()
```

You look at what an example of this repetition looks like for a single person.

```{r uniqueness-example}
cohort_clean |>
  filter(person_id == repeated_entries$person_id[1]) |>
  select(record_id, person_id, visit_date, weight_kg) |>
  arrange(visit_date)
```

Two fixes are needed. Identical rows are one submission sent twice, and
`distinct()` handles them cleanly. The rest are the same child on the same day
with a **new** `record_id` and a different weight. This is double data entry
where the attempts disagree. You keep the newer one, identified here by the
larger `record_id`, on the assumption that a value gets re-keyed because someone
noticed a problem with the first. As before, document the assumption.

```{r uniqueness-fix}
# TODO: drop exact duplicate rows, then for same-day repeats of the same
# person, keep only the row with the largest record_id.
cohort_clean <- cohort_clean |>
  ___() |>
  group_by(person_id, visit_date) |>
  ___(record_id, n = 1, with_ties = FALSE) |>
  ungroup()

nrow(cohort_clean)
```

---

### 6. Timeliness

To assess timeliness, you check whether data arrived, and was collected,
quickly enough to be useful.

The bureau tells you that they received data from most woredas in January, but
that Jimma Town only provided data in March. You calculate the reporting lag
(the gap in days between `visit_date` and `entry_date`) per woreda, and look at
the five woredas with the longest median lag.

```{r timeliness-diagnose}
cohort_clean |>
  mutate(lag_days = as.numeric(entry_date - visit_date)) |>
  group_by(woreda_std) |>
  summarise(median_lag = median(lag_days, na.rm = TRUE), .groups = "drop") |>
  arrange(desc(median_lag)) |>
  head(5)
```

**Question:** How does Jimma Town's reporting lag compare to the other woredas above? Does
that match what the bureau told you, and what might explain it?

Since there is a reporting lag between a visit happening and it being entered,
you check the last month for which you have visit dates against the months
before it, in case the most recent month has not caught up yet.

```{r timeliness-month-diagnose}
cohort_clean |>
  mutate(month = floor_date(visit_date, "month")) |>
  count(month) |>
  arrange(month) |>
  tail(4)
```

**Question:** How does the final month's record count compare to the months before it? Given
that, and the reporting lag you just saw, should you keep it as-is?

```{r timeliness-fix}
# TODO: drop the final (most recent) month - keep only visits from before
# the maximum month present in the data.
cohort_clean <- cohort_clean |>
  mutate(month = floor_date(visit_date, "month")) |>
  filter(___)

nrow(cohort_clean)
```

---

### Document your pipeline

In this part you assessed each data dimension and fixed issues as they arose.
Another approach is to collect every fix into a single pipeline that cleans
everything in one place, which is easier to audit, rerun, and hand over to
someone else.

**Optional challenge:** once every fix above is working, try rewriting them as a single piped
pipeline starting from `read_csv("data/nutrition_screening.csv")`, so the whole cleaning process
runs in one block.

Whichever approach you take, write down what you did and assumed for each of the six dimensions
above (completeness, accuracy, validity, consistency, uniqueness, timeliness).

---

## Part 2 - Working with shapefiles

In this part you focus on working with shapefiles and joining your various
datasets to them. You will use two shapefiles. The first is supplied by the
regional bureau (`woredas.shp`) and contains the 56 woredas they sampled. The
second is `eth.shp`, which you downloaded yourself from the public national
database for country-wide context.

#### `woredas.shp` - 56 sampled woredas, UTM 37N (EPSG:32637)

| Field | Type | Description |
|---|---|---|
| `wor_pcode` | text | Woreda code |
| `wor_name` | text | Woreda name |
| `geometry` | MULTIPOLYGON | The woreda's boundary shape, in UTM 37N. |

```{r woredas-shape}
woredas <- st_read("data/woredas.shp", quiet = TRUE)
```

#### `eth.shp` - 1,148 national woredas, lon/lat (EPSG:4326)

Source: [HDX](https://data.humdata.org/dataset/cod-ab-eth). Woreda codes and
names match `woredas.shp`, and each woreda also carries the name and code of its
zone and region. Note that Oromia's zones have changed since the rainfall data was produced, with some
zones splitting. The shapefile has been edited to merge those zones back
together, so `zone_pcode` matches all 14 zones in the rainfall file.

| Field | Type | Description |
|---|---|---|
| `wor_pcode` | text | Woreda code |
| `wor_name` | text | Woreda name |
| `zone_pcode` | text | Zone code |
| `zone_name` | text | Zone name |
| `reg_pcode` | text | Region code |
| `reg_name` | text | Region name |
| `geometry` | MULTIPOLYGON | The woreda's boundary shape |

```{r eth-shape}
eth <- st_read("data/eth.shp", quiet = TRUE)
```

### 1. What is in a shapefile?

You start by printing the object and reading its header.

```{r what-is-sf}
woredas
```

The header is the useful part. **Geometry type** is MULTIPOLYGON, so some
woredas are in more than one piece. The **bounding box** numbers are in the
hundreds of thousands, so these are metres, not degrees. The **CRS** is UTM
zone 37N.

There are two things to watch for when you subset or summarise an sf object.

```{r sticky-geometry}
woredas |> select(wor_name)

eth |> st_drop_geometry() |> count(zone_name) |> head()
```

`select(wor_name)` returns two columns, not one, because geometry is **sticky**
and comes along unless you explicitly drop it. That is usually what you want on
a map and never what you want in a summary table. `woredas.shp` only has 56
rows, too few for that to cost you anything either way, so the second example
uses `eth`. 1,148 rows is where `st_drop_geometry()` first actually saves you
something, and where forgetting to drop it before a `group_by()` can turn
seconds into minutes.

---

### 2. Where is your data?

You add a flag to `eth` marking which woredas were sampled, so that you can see
your woredas in context.

```{r eth-enrich}
# TODO: mark which of eth's woredas are among the 56 sampled ones in
# `woredas` (match on wor_pcode)
eth <- eth |>
  mutate(sampled = ___)

eth |>
  st_drop_geometry() |>
  count(sampled)
```

This is an ordinary `mutate()` against a vector, not a join in the sf sense. It
matches on `wor_pcode`, a shared code, so it works even though `eth` is lon/lat
and `woredas` is UTM. CRS only matters once two geometries interact directly, as
Task 3 shows.

`eth` now carries everything you need: the national picture, the admin
hierarchy, and a flag for the 56 sampled woredas. You use `woredas` in Task 3
because it is the file with the wrong projection, but all other tasks use
`eth`, filtered to `sampled` when you mean just those 56.

Carrying the full country through every step costs more than it needs to.
Working from `woredas` and joining back to `eth` only at the end would be
faster, but keeping one object throughout is easier to follow.

You now plot some maps and look at the woredas your data comes from. You plot
the following five maps:

1. All woredas.

```{r plot-country, fig.height=5}
ggplot(eth) +
  geom_sf(fill = "gray90", colour = "black", linewidth = 0.1) +
  labs(title = "All woredas") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5))
```

2. Colour the woredas that are in Oromia.

```{r plot-region, fig.height=5}
ggplot(eth) +
  # TODO: TRUE for woredas in the Oromia region (reg_name)
  geom_sf(aes(fill = ), colour = "black", linewidth = 0.1) +
  scale_fill_manual(values = c("TRUE" = "#4A90D9", "FALSE" = "gray90")) +
  labs(title = "Woredas in Oromia", fill = "Oromia") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5))
```

3. Colour the woredas that were sampled.

```{r plot-sampled, fig.height=5}
ggplot(eth) +
  # TODO: geom_sf, filled by the sampled flag you created above
  ___
  scale_fill_manual(values = c("TRUE" = "#4A90D9", "FALSE" = "gray90")) +
  labs(title = "Sampled woredas", fill = "Sampled") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5))
```

4. Zoom into the Oromia region and plot the woredas that were sampled.

```{r plot-zoom, fig.height=5}
bb2 <- st_bbox(filter(eth, reg_name == "Oromia"))

ggplot(eth) +
  geom_sf(aes(fill = sampled), colour = "black", linewidth = 0.1) +
  scale_fill_manual(values = c("TRUE" = "#4A90D9", "FALSE" = "gray90")) +
  # TODO: xlim needs bb2's x min/x max, ylim needs its y min/y max
  # print bb2 to see the named values it holds and select the elements using
  # $
  coord_sf(xlim = c(___, ___), ylim = c(___, ___)) +
  labs(title = "Sampled woredas, zoomed to Oromia", fill = "Sampled) +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5))
```

5. Plot only the sampled woredas and nothing else.

```{r plot-sampled-only, fig.height=5}
# TODO: keep only the sampled woredas
ggplot(filter(eth, ___)) +
  geom_sf(fill = "#4A90D9", colour = "gray90") +
  labs(title = "Sampled woredas only") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5))
```

**Question:** Compare plot 4 and plot 5. What's the difference between how `coord_sf()` and
`filter()` handle the woredas outside your sample?

**You need to choose a plot to show the sampled woredas in a report. Which would you choose?**

---

### 3. Coordinate reference systems

#### `health_posts.csv` - one row per health post

Locations of the health posts that captured data in the screening register
(`nutrition_screening.csv`).

| Field | Type | Description |
|---|---|---|
| `health_post` | text | Identifier. Joins to `nutrition_screening.csv`. |
| `hp_name` | text | Human-readable name. |
| `lon` | number | Longitude. A handful of rows have longitude and latitude swapped; see Part 2, Task 4. |
| `lat` | number | Latitude. A few rows are missing entirely; one sits exactly on a woreda boundary, by design. |

```{r hps}
posts <- read_csv("data/health_posts.csv")
```

You want to join the posts to one of your shapefiles so that you can see where
the health posts that collected your data are. You receive a csv with lon/lat
details, and you convert it into an `sf` object - geometry does not always have
to come from a shapefile.

```{r crs-check}
# 4326 = EPSG code for WGS84 (plain lon/lat)
posts_sf <- st_as_sf(posts, coords = c("lon", "lat"), crs = 4326)

posts_sf
```

Note that the geometry of `posts_sf` is a POINT.

Now you try to join `posts_sf` and `woredas`, which have different CRSs.

```{r crs-fail, error=TRUE}
st_join(posts_sf, woredas, join = st_within)
```

`sf` refuses: `st_crs(x) == st_crs(y) is not TRUE`. Metres and degrees are not
interchangeable, and a silent wrong answer would be worse than an error - every
point would land in the Gulf of Guinea, near (0, 0).

You transform the woredas into lon/lat so that both objects share a CRS.

```{r crs-fix}
# TODO: transform woredas into lon/lat (EPSG:4326) so it shares a CRS with
# posts_sf
woredas_ll <- st_transform(woredas, crs=___)
st_crs(woredas_ll)$input
```

> **Note `st_transform()` vs `st_set_crs()`.** Transform recalculates every
> coordinate into the new system. Set just writes a different label onto
> unchanged numbers - appropriate only when the CRS was missing or wrong in the
> first place. Using `st_set_crs()` where `st_transform()` was meant moves
> the data thousands of kilometres and produces no error at all.

---

### 4. Point in polygon

You now join the posts to `woredas_ll` to find which polygon each health post
falls within. `st_join()` needs a predicate to decide what counts as a match.
You use `st_within`, which only counts a point as inside a polygon if it is
fully enclosed by it. The common alternative is `st_intersects`, which is more
generous: it counts anything touching the polygon at all, including its edge.

```{r spatial-join}
# TODO: join posts_sf to woredas_ll, keeping one row per post. Use the
# stricter "fully enclosed" st_within.
joined <- st_join(posts_sf, woredas_ll, join = ___)

c(total_posts = nrow(joined),
  posts_matched = sum(!is.na(joined$wor_pcode)),
  posts_unmatched = sum(is.na(joined$wor_pcode)))
```

The order of the arguments matters: `st_join()` keeps the first object's rows and
geometry, so `posts_sf` has to come first for the result to have one row per
post. Swapping them gives one row per woreda instead, and
`st_within(woredas_ll, posts_sf)` would ask whether each polygon sits inside a
point.

Some posts were not matched. You plot the posts to see whether you can work out
why.

```{r plot-failures, fig.height=5}
ggplot() +
  geom_sf(data = eth, fill = "grey95", colour = "black", linewidth = 0.1) +
  geom_sf(data = woredas, fill = "blue", colour = "black") +
  geom_sf(data = st_jitter(joined, amount = 0.3), colour = "#D95F02", size = 1.5) +
  coord_sf(xlim = c(0, 46), ylim = c(0, 45)) +
  labs(title = "Health posts that landed nowhere", caption = sim_caption) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

Next you look at the longitude and latitude of the problematic posts.

```{r inspect-failures}
joined |>
  filter(is.na(wor_pcode)) |>
  # st_coordinates() pulls the X/Y values out of a geometry column into a
  # plain matrix - column 1 is X (lon), column 2 is Y (lat)
  mutate(lon = st_coordinates(geometry)[, 1],
         lat = st_coordinates(geometry)[, 2]) |>
  st_drop_geometry() |>
  select(health_post, lon, lat)

# Ethiopia's bounding box, for comparison
bb <- st_bbox(eth)
bb
```

**Question:** Compare the coordinates of the unmatched posts to Ethiopia's bounding box above.
What do you think has gone wrong with these rows?

You flag the swapped rows, swap them back, and redo the join.

```{r fix-swapped}
posts_fixed <- posts |>
  mutate(
    # TODO: flag rows where lon/lat look swapped for Ethiopia (i.e. lon is
    # suspiciously small and lat is suspiciously large), then swap them back
    swapped = ___,
    lon_fix = if_else(swapped, lat, lon),
    lat_fix = if_else(swapped, lon, lat)
  )

posts_sf2 <- posts_fixed |>
  st_as_sf(coords = c("lon_fix", "lat_fix"), crs = 4326)

joined2 <- st_join(posts_sf2, eth, join = st_within)

c(posts_unmatched = sum(is.na(joined2$wor_pcode)))
```

> **What if a post sits exactly on a woreda boundary?**

---

### 5. Joining tables to a shapefile

#### `woreda_population.csv` - one row per woreda per year (2019-2024)

The number of people, and the number of under-5s, living in each woreda each
year.

| Field | Type | Description |
|---|---|---|
| `woreda_pcode` | text | Woreda code |
| `woreda_name` | text | Woreda name |
| `year` | number | Calendar year of the projection |
| `total_pop` | number | Total projected population |
| `pop_under2` | number | Projected population under 2 years |
| `pop_under5` | number | Projected population under 5 years |

```{r wor_pop}
pop <- read_csv("data/woreda_population.csv")
```

You can also join a table to a shapefile as normal, as long as there is a
column to join on.

```{r pop-join}
before <- nrow(woredas)
# TODO: join on the woreda code - match woredas' code field to pop's
# woreda_pcode
wor_pop <- woredas |> left_join(pop, by = c(___ = "woreda_pcode"))

c(before = before,
  pop_years = n_distinct(pop$year),
  after = nrow(wor_pop))
```

The number of rows has grown, since each woreda now maps to each population year
(that is, the number of rows before multiplied by the number of population
years). The geometry is repeated across these rows, so exercise caution when
joining to shapefiles, as the output can grow quickly.

Now you try making the same join using woreda names.

```{r name-join}
# TODO: join on the woreda name instead - match pop's equivalent name field
by_name <- woredas |>
  left_join(pop, by = c("wor_name" = ___))

sum(is.na(by_name$total_pop))

by_name |> filter(is.na(total_pop)) |> st_drop_geometry() |> select(wor_name)
```

**Question:** Look at the woredas that failed to match by name above, and compare their
spelling to the equivalent entries in `woreda_population.csv`. What's going wrong, and what does
this tell you about joining on names versus codes?

---

### 6. Aggregating up

#### `zone_rainfall.csv` - one row per zone per month

Monthly rainfall for each zone, already summarised from satellite data.

| Field | Type | Description |
|---|---|---|
| `zone_pcode` | text | Zone unique identifier |
| `date` | date | First of the month that the rainfall applies to |
| `n_pixels` | number | Climate Hazards Group InfraRed Precipitation with Station data (CHIRPS) pixels averaged for this zone |
| `rain_mm` | number | Rainfall total for the month, in mm. |
| `rain_lta_mm` | number | Long-term average rainfall for that same calendar month, in mm |
| `rain_anom_pct` | number | `rain_mm` as a **percentage of `rain_lta_mm`**. 100 is normal, below 100 is drier than usual, above 100 is wetter |

```{r rainfall}
rainfall <- read_csv("data/zone_rainfall.csv")
```

Rainfall only exists at a zone level. You can either assume that each woreda
experiences the same amount of rain as the entire zone, or aggregate the
woredas to a zone level. Given that you need to answer policy questions related
to zones, you aggregate. You also keep track of which zones were sampled,
relative to the other zones.

```{r zones-grouped, fig.height=4}
# TODO: group by the zone code and zone name to dissolve woredas up to zone
# level, then flag a zone as sampled if any of its woredas were sampled
zones <- eth |>
  group_by(___, ___) |>
  summarise(sampled = ifelse(any(sampled), TRUE, FALSE),
            .groups = "drop")

nrow(zones)

ggplot(zones) +
  geom_sf(aes(fill = sampled), colour = "black", linewidth = 0.3, show.legend = FALSE) +
  labs(title = "Dissolved",
       subtitle = "14 zone shapes") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))
```

Note that `group_by() |> summarise()` on an sf object goes a step
further: it unions the geometries that share a group.

Now you merge the zones to the rainfall data and plot a single month.

```{r attach-rainfall, fig.height=5}
zone_rain <- zones |>
  left_join(rainfall, by = "zone_pcode")

# Rainfall for 2022-06-01
ggplot() +
  geom_sf(data = eth, fill = "gray90", colour = "black", linewidth = 0.2) +
  geom_sf(data = zone_rain |> filter(date == as.Date("2022-06-01")), 
  aes(fill = rain_anom_pct), colour = "black", linewidth = 0.3) +
  scale_fill_viridis_c() +
  labs(title = "Rainfall, June 2022",
       subtitle = "Zone shapes are the dissolved sample, not the true extent",
       fill = "% of normal") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))
```

---

### 7. Tying it all together

You now need to tie the previous steps together and create a single object that
carries the screening records along with population and rainfall data.

You start with the screening records. You can either join them to `eth` at
woreda level on `woreda_name`, or you can go via the health post coordinates.
Joining on coordinates is more reliable than a name, and lets you reuse
`posts_sf2` from Part 2, Task 4 - the version with the swapped lon/lat rows
corrected and any post missing coordinates entirely dropped. To join using
coordinates, you first join to the health post object, then use `st_within()`
to place each record in a zone.

```{r attach-geography}
cohort <- cohort_clean

cohort_geo <- cohort |>
  left_join(posts_sf2 |> select(health_post), by = "health_post") |>
  # TODO: the join above keeps the geometry column but drops the sf class -
  # restore it so st_within() has something to work with
  ___

cohort_geo_before <- nrow(cohort_geo)

combined_dataset <- cohort_geo |>
  st_join(eth |>
   select(wor_pcode, zone_pcode, zone_name, reg_pcode, reg_name),
   join = st_within) |>
  st_drop_geometry()

c(before = nrow(cohort_geo), after = nrow(combined_dataset))
```

`cohort` is a plain data frame, so joining it to the sf object `posts_sf2` keeps
the geometry column but not the `sf` class. `st_as_sf()` restores it. Because
`st_within()` is strict about boundaries, you check the row counts to confirm
that no points on a zone edge were dropped.

Finally, you attach population and rainfall to each visit.

```{r cohort-full}
rainfall <- rainfall |> rename(month = date)

# TODO: join population by woreda code + year, and rainfall by zone code +
# month - population is per woreda per year, rainfall is per zone per month.
combined_dataset <- combined_dataset |>
  mutate(year = year(visit_date)) |>
  left_join(pop |> select(-woreda_name), by = ___) |>
  left_join(rainfall |> select(-n_pixels), by = ___)

nrow(combined_dataset)
```

---

In this part you learnt how to work with shapefiles and join data to them.
Using what you have learned, you created `combined_dataset`, which you use in
the next part. It has one row per screening visit, with the rainfall for that
visit's zone that month and the population for its woreda that year attached to
each row.

**What this part has not covered:** small polygons a coarse grid misses
entirely, bounding boxes over-claiming on curved units, polygon-to-polygon
overlap and area weighting, and rasters. A tutorial covering these is in your
materials. When a spatial operation errors, try `st_is_valid()`,
`st_make_valid()`, `sf_use_s2(FALSE)` and `st_cast()` before anything else.

---

## Part 3 - Building an outcome, and looking at it

### 1. From a weight to an outcome

To answer your policy question you need to know who is underweight. Raw weight
alone cannot tell you that. A 12 kg two-year-old and a 12 kg four-year-old are
not equally nourished, so weight only means something once it is compared
against what is expected for a child of that age and sex.

WHO's Child Growth Standards supply that reference, and the `anthro` package
turns each child's raw weight into a **weight-for-age z-score (WAZ)**. A WAZ
tells you how many standard deviations a child sits from the median child of
the same age and sex. According to WHO cutoffs, below -2 is underweight and
below -3 is severe.

`anthro_zscores()` is particular about its inputs. Sex must be coded `1`/`2`
rather than `"M"`/`"F"`, age must be in whole months, and a `measure` argument
tells it how height was taken. Children under 24 months are measured lying down
(`l`), older children standing up (`h`); the two give different numbers for the
same child, which the package adjusts for.

```{r zscores}
z <- anthro_zscores(
  sex = if_else(combined_dataset$sex_std == "M", 1, 2),
  age = combined_dataset$age_months,
  is_age_in_month = TRUE,
  weight = combined_dataset$weight_kg,
  lenhei = combined_dataset$height_cm,
  measure = if_else(combined_dataset$age_months < 24, "l", "h")
)

combined_dataset <- combined_dataset |>
  mutate(waz = z$zwei)

combined_dataset |>
  summarise(
    children = n(),
    no_zscore = sum(is.na(waz))
  )
```

---

### 2. Look at the outcome before you use it

Now look at a histogram of the outcome variable you have created, with dashed
lines at -2 and -3, the WHO thresholds for underweight and severely underweight
respectively.

```{r hist, fig.height=4}
combined_dataset |>
  ggplot(aes(waz)) +
  geom_histogram(bins = 60, fill = "#4A90D9") +
  geom_vline(xintercept = c(-3, -2), linetype = "dashed", colour = "#D95F02") +
  labs(title = "Weight-for-age z-scores",
       subtitle = "Dashed lines: underweight (-2) and severe (-3)",
       x = "WAZ", y = "Children", caption = sim_caption) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        panel.border = element_blank(),
        axis.line = element_line(colour = "black"))
```

**Question:** Looking at the histogram above, what does the shape of the distribution tell you,
especially relative to the -2/-3 cutoffs? Next you create a boxplot that splits the same
distribution by zone.

```{r box, fig.height=4}
combined_dataset |>
  ggplot(aes(x = fct_reorder(zone_name, waz), y = waz)) +
  geom_boxplot(fill = "#4A90D9", alpha = 0.6, outlier.alpha = 0.2) +
  geom_hline(yintercept = c(-3, -2), linetype = "dashed", colour = "#D95F02") +
  coord_flip() +
  labs(title = "Weight-for-age by zone", x = NULL, y = "WAZ",
       caption = sim_caption) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        panel.border = element_blank(),
        axis.line = element_line(colour = "black"))
```

**Question:** Compare the boxplots across zones. Is the left-skew you saw in the histogram
common to every zone, or driven by a few? Do any zones stand out from the rest?

---

### 3. Mapping the children underweight

While the z-score is useful, you are more interested in the number of children
who are underweight and who are severely underweight.

```{r underweight-flags}
# TODO: using the WHO cutoffs from Part 3.1 (-2 underweight, -3 severe),
# flag each child as underweight, severe, or total_underweight (either one)
combined_dataset <- combined_dataset |>
  mutate(
    underweight = ___,
    severe = ___,
    total_underweight = underweight | severe
  )

combined_dataset |>
  summarise(
    pct_underweight = round(100 * mean(underweight), 1),
    pct_severe = round(100 * mean(severe), 1),
    pct_total_underweight = round(100 * mean(total_underweight), 1)
  )
```

Now you plot your data. You consider three views: the number of children
screened, the number underweight, and the proportion of screened children who
are underweight. First you build the zone-level summary those maps need.

```{r by-zone}
# pop_under5 is a per-woreda figure, repeated on every child's visit row in
# combined_dataset - summing it directly would add that same number in once
# per visit instead of once per woreda, so take distinct woreda values first.
zone_pop <- combined_dataset |>
  distinct(wor_pcode, zone_pcode, pop_under5) |>
  group_by(zone_pcode) |>
  summarise(pop_under5 = sum(pop_under5, na.rm = TRUE), .groups = "drop")

zone_summary <- combined_dataset |>
  group_by(zone_pcode) |>
  summarise(
    n_screened = n(),
    n_underweight = sum(underweight),
    n_severe = sum(severe),
    n_total_underweight = sum(total_underweight),
    # TODO: proportion of screened children who are underweight, 
    # use n_total_underweight as the denominator
    prop_screened_underweight = ___,
    .groups = "drop"
  ) |>
  left_join(zone_pop, by = "zone_pcode")

zone_data <- zones |> left_join(zone_summary, by = "zone_pcode")
```

You define a helper function to speed up the plotting, then use it to create
three maps.

```{r three-maps, fig.width=7, fig.height=15, warning=FALSE}
p <- function(var, title, fill_title, labels = scales::label_number()) {
  bb <- st_bbox(filter(zone_data, sampled == TRUE))

  ggplot() +
    geom_sf(data = zone_data, aes(fill = .data[[var]]), colour = "black", linewidth = 0.2) +
    geom_sf_label(data = filter(zone_data, sampled), aes(label = zone_name), size = 2,
                  linewidth = 0, fill = scales::alpha("gray90", 0.7)) +
    # to zoom in, uncomment this line
    # coord_sf(xlim = bb[c("xmin", "xmax")], ylim = bb[c("ymin", "ymax")]) +
    scale_fill_viridis_c(direction = -1, na.value = "grey95", labels = labels) +
    labs(title = title, fill = fill_title, caption = sim_caption) +
    theme_void() +
    theme(plot.title = element_text(hjust = 0.5))
}

# TODO: fill in each blank with the column to map, as a quoted string (e.g.
# "n_severe") 
p(___ , "a) Underweight children", "Children") +
p(___ , "b) Children screened", "Children") +
p(___ , "c) Proportion underweight", "Proportion",
  labels = scales::label_percent()) +
  plot_layout(ncol = 1)
```

**Question:** Compare Ilu Aba Bora and Arsi across the three maps above. What do you notice?

Colour on a map is hard to rank precisely. The same children as bars, sorted and
split by severity, make the point more clearly:

```{r severity-bars, fig.height=9}
zone_severity <- zone_data |>
  st_drop_geometry() |>
  filter(sampled) |>
  select(zone_name, n_screened, Underweight = n_underweight, Severe = n_severe) |>
  pivot_longer(c(Underweight, Severe), names_to = "severity", values_to = "n") |>
  mutate(
    severity = factor(severity, levels = c("Underweight", "Severe")),
    prop_screened = n / n_screened
  )

q <- function(value_var, title, y_lab, labels = scales::label_number()) {
  ggplot(zone_severity, aes(x = fct_reorder(zone_name, .data[[value_var]], sum),
                            y = .data[[value_var]], fill = severity)) +
    geom_col() +
    coord_flip() +
    scale_y_continuous(labels = labels) +
    scale_fill_manual(values = c(Underweight = "#4A90D9", Severe = "#D95F02")) +
    labs(title = title, x = NULL, y = y_lab, fill = NULL, caption = sim_caption) +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5),
          panel.border = element_blank(),
          axis.line = element_line(colour = "black"))
}

# TODO: fill in each blank with the column to plot, as a quoted string (e.g.
# "n_severe") 
q(___, "Underweight children",
  "Children") +
q(___, "Proportion of children screened",
  "Proportion", labels = scales::label_percent()) +
  plot_layout(ncol = 1)
```

---

### 4. Coverage

Map (c) ranks zones by the proportion underweight among children screened.
But that proportion only describes the children who came, not the children who
live there. To take this into account you need to investigate coverage. You add
coverage to `zone_data` and map it alongside the under-5 population.

```{r coverage, fig.width=7, fig.height=10}
# TODO: coverage = children screened as a proportion of the under-5 population
zone_data <- zone_data |>
  mutate(coverage = ___)

p("pop_under5", "d) Under-5 population",
  "Children under-5") +
p("coverage", "e) Coverage (screened ÷ under-5 population)",
  "Screened", labels = scales::label_percent(accuracy = 0.1)) +
  plot_layout(ncol = 1)
```

**Question:** Looking at maps (d) and (e), how does coverage vary across zones?
 Which zone has the highest coverage, and which the lowest?

Ranking the zones by the proportion underweight, alongside their coverage,
gives:

```{r final-ranking}
zone_data |>
  st_drop_geometry() |>
  filter(sampled) |>
  arrange(desc(prop_screened_underweight)) |>
  transmute(
    zone_name,
    n_screened,
    n_total_underweight,
    # TODO: proportion of screened children who are underweight,
    #  use n_total_underweight as the denominator
    pct_underweight = round(100 * ___, 1),
    pct_coverage = round(100 * coverage, 2)
  )
```

**So, which zones in Oromia have the highest malnutrition, based on the ranking above? 
What limitations should you flag before reporting this?**

---

### 5. Over time

Now you need to address question 2. As a reminder, `rain_anom_pct` is that
month's rainfall as a percentage of the long-term average for that same month
in that same zone: 100 is normal, below 100 is drier than usual, above 100 is
wetter. This is the same idea as the proportions above. Absolute rainfall is not
comparable across seasons or zones, so it needs a denominator too, and here the
denominator is what that zone normally gets at that time of year.

Start by looking at the proportion underweight and `rain_anom_pct` overall.

```{r time-plot, fig.height=5}
monthly <- combined_dataset |>
  group_by(month) |>
  summarise(n_screened = n(),
            n_total_underweight = sum(total_underweight),
            prop = n_total_underweight / n_screened,
            rain_pct = mean(rain_anom_pct, na.rm = TRUE),
            .groups = "drop")

# TODO: plot rainfall over time - map x to the time variable you grouped by
# and y to the rainfall variable you computed above.
p_rain <- ggplot(monthly, aes(x=___, y=___)) +
  geom_line(colour = "#4A90D9", linewidth = 0.9) +
  labs(title = "Nutrition follows the rains, with a delay",
       x = NULL, y = "Rainfall, % of normal") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        panel.border = element_blank(),
        axis.line = element_line(colour = "black"))

# TODO: plot proportion underweight over time - map x to the same time
# variable as p_rain, and y to the proportion you computed above.
p_prop <- ggplot(monthly, aes(x=___, y=___)) +
  geom_line(colour = "#D95F02", linewidth = 0.9) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = NULL, y = "Proportion underweight", caption = sim_caption) +
  theme_bw() +
  theme(panel.border = element_blank(),
        axis.line = element_line(colour = "black"))

p_rain / p_prop
```

**Question:** Look at the two panels above. Does the proportion underweight move at the same
time as rainfall, or with a delay?

One explanation for a lag like this is the chain from poor rains to crop and livestock losses,
then to household food insecurity, and finally to malnourishment. It is important to be clear,
though, that nothing here establishes that chain. Two series moving in sequence is consistent
with it, but it may also support many other explanations.

**How long is that delay?** Rather than guess, you check every lag from 0 to 9
months and see which one actually correlates with the proportion underweight.

```{r lag-scan, fig.height=4}
lag_scan <- data.frame(lag = 0:9) |>
  mutate(correlation = sapply(lag, function(k) {
    cor(monthly$prop, lag(monthly$rain_pct, k), use = "complete.obs")
  }))

ggplot(lag_scan, aes(lag, correlation)) +
  geom_col(fill = "#4A90D9") +
  geom_hline(yintercept = 0, colour = "black") +
  scale_x_continuous(breaks = 0:9) +
  labs(title = "How many months back does rainfall predict underweight?",
       x = "Lag (months)", y = "Correlation with proportion underweight",
       caption = sim_caption) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        panel.border = element_blank(),
        axis.line = element_line(colour = "black"))
```

**Question:** Looking at the bar chart above, over which range of lags does rainfall most
strongly correlate with the proportion underweight? From here you use the average across that
window, `rain_lag`, rather than picking a single month out of it.

Based on the window you identified above, you plot the averaged, lagged rainfall and overlay the
two series so that they move together. The relationship should be clearer now.

```{r lag-overlay, fig.height=4.5}
# TODO: use the lag window you identified from the chart above - average
# together lag(rain_pct, k) for each month k in that window starting at 3
monthly_shifted <- monthly |>
  mutate(rain_lag_pct = (lag(rain_pct, ___) + lag(rain_pct, ___) +
                          lag(rain_pct, ___) + lag(rain_pct, ___)) / 4)

scale_factor <- max(monthly_shifted$rain_lag_pct, na.rm = TRUE) /
  max(monthly_shifted$prop, na.rm = TRUE)

ggplot(monthly_shifted, aes(month)) +
  geom_line(aes(y = rain_lag_pct), colour = "#4A90D9", linewidth = 0.9, na.rm = TRUE) +
  geom_line(aes(y = prop * scale_factor), colour = "#D95F02", linewidth = 0.9) +
  scale_y_continuous(
    name = "Rainfall 3-6 months earlier (% of normal)",
    sec.axis = sec_axis(~ . / scale_factor, name = "Proportion underweight",
                         labels = scales::percent)
  ) +
  labs(title = "Shifted by the lag, the two lines move together",
       subtitle = "Blue: rainfall 3-6 months earlier (left axis). Orange: proportion underweight (right axis).",
       x = NULL, caption = sim_caption) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        axis.title.y.left = element_text(colour = "#4A90D9"),
        axis.title.y.right = element_text(colour = "#D95F02"),
        panel.border = element_blank(),
        axis.line.x.bottom = element_line(colour = "black"),
        axis.line.y.left = element_line(colour = "black"),
        axis.line.y.right = element_line(colour = "black"))
```

---

### 6. The relationship

Now that you know there is a relationship overall, look at it at a zone level.
You start by creating a `zone_month` object, which is similar to `monthly` but
also groups by `zone_pcode` and `zone_name`.

```{r zone-month}
zone_month <- combined_dataset |>
  group_by(zone_pcode, zone_name, month) |>
  summarise(n_screened = n(),
            n_total_underweight = sum(total_underweight),
            prop = n_total_underweight / n_screened,
            rain_anom_pct = mean(rain_anom_pct),
            .groups = "drop")

# TODO: use the lag window you identified from the chart above - average
# together lag(rain_pct, k) for each month k in that window starting at 3
zone_month <- zone_month |>
  arrange(zone_pcode, month) |>
  group_by(zone_pcode) |>
  mutate(rain_lag = (lag(rain_anom_pct, ___) + lag(rain_anom_pct, ___) +
                     lag(rain_anom_pct, ___) + lag(rain_anom_pct, ___)) / 4) |>
  ungroup() |>
  filter(!is.na(rain_lag)) # lagging costs you the first six months of data
```

You start by looking at a scatter of lagged rainfall against the proportion
underweight.

```{r scatter, fig.height=4}
zone_month |>
  filter(n_screened >= 25) |>
  # TODO: map x to the lagged rainfall column and y to the proportion
  # underweight column, both created above.
  ggplot(aes(x=___, y=___)) +
  geom_point(aes(), alpha = 0.25, colour = "#4A90D9") +
  geom_vline(xintercept = 100, linetype = "dotted", colour = "black") +
  geom_smooth(aes(weight = n_screened), method = "glm",
              method.args = list(family = quasibinomial()),
              se = FALSE, colour = "#D95F02") +
  labs(title = "Underweight against the previous season's rainfall",
       subtitle = "Each point is a zone-month; below 100% is drier than normal",
       x = "Rainfall 3-6 months earlier (% of normal)",
       y = "Proportion underweight", caption = sim_caption) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        panel.border = element_blank(),
        axis.line = element_line(colour = "black"))
```

**Question:** Looking at the scatter plot and fitted curve above, what's the relationship
between lagged rainfall and the proportion of children underweight? Does it look the same across
the whole range of rainfall values, or does it change?

**Is that relationship stable, or is it one drought year doing all the work?**
To answer this, look at the correlation within each year.

```{r yearly-corr, fig.height=4}
yearly_corr <- zone_month |>
  mutate(year = year(month)) |>
  group_by(year) |>
  # Neither prop nor rain_lag has NAs left at this point, so a plain cor() is
  # safe here. If either one did, cor()'s use = "complete.obs" argument would
  # drop incomplete pairs instead of returning NA.
  summarise(
    # TODO: correlation between prop and rain_lag for this year, use the cor function
    correlation = ___,
    n_obs = sum(!is.na(prop) & !is.na(rain_lag)),
    .groups = "drop"
  )

# TODO: bar per year - x is the year column (wrapped in factor() so it's
# treated as discrete), y is the correlation you computed above.
ggplot(yearly_corr, aes(factor(___), ___)) +
  geom_col(fill = "#4A90D9") +
  geom_hline(yintercept = 0, colour = "black") +
  geom_text(aes(label = paste0("n=", n_obs),
                y = pmin(correlation, 0) - 0.03),
            size = 3, colour = "grey30") +
  labs(title = "The rain-underweight correlation, year by year",
       subtitle = "n = zone-month observations feeding each year's correlation",
       x = NULL, y = "Correlation (prop underweight vs. 3-6 month rain lag)",
       caption = sim_caption) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        panel.border = element_blank(),
        axis.line = element_line(colour = "black"))
```

**Is it the same in every zone, or concentrated in a few?**

```{r zone-corr-map, fig.height=5, warning=FALSE}
zone_corr <- zone_month |>
  group_by(zone_pcode) |>
  # Same as above: a plain cor() is safe since prop and rain_lag are already
  # complete, but use = "complete.obs" is what you'd reach for if either had NAs.
  summarise(n_months = n(),
            # TODO: correlation between prop and rain_lag for this zone
            correlation = ___,
            .groups = "drop")

zone_corr_map <- zones |>
  left_join(zone_corr, by = "zone_pcode") |>
  filter(!is.na(correlation))

# Zoom to the zones that actually carry a correlation. `bb` further up is
# Ethiopia's national bounding box and would zoom out to the whole country.
bb_corr <- st_bbox(zone_corr_map)

ggplot() +
  geom_sf(data = zones, fill = "grey95", colour = "black", linewidth = 0.1) +
  geom_sf(data = zone_corr_map, aes(fill = correlation), colour = "black", linewidth = 0.2) +
  geom_sf_label(data = zone_corr_map, aes(label = zone_name), size = 2,
                linewidth = 0, fill = scales::alpha("gray90", 0.7)) +
  scale_fill_viridis_c(limits = c(-0.6, -1)) +
  coord_sf(xlim = bb_corr[c("xmin", "xmax")], ylim = bb_corr[c("ymin", "ymax")]) +
  labs(title = "How rain-sensitive is each zone?",
       subtitle = "Pearson correlation between lagged rainfall and proportion underweight, pooled across all six years",
       fill = "Pearson correlation", caption = sim_caption) +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))
```

Note that a correlation describes how two series move together. It says nothing
about how many children are underweight in a zone, so these values cannot be
used to rank zones as better or worse off. The two are separate, and reading them together is
what makes the zone comparisons informative.

**So, is rainfall associated with child malnutrition in Oromia?**

---

## Before you leave

With that, you have enough to send the Regional Health Bureau a preliminary
reply. They asked two questions:

1. **Which zones in Oromia have the highest malnutrition?**
2. **Is rainfall associated with child malnutrition in Oromia?**

Write your own answers to both, and note what limitations you'd flag before this goes to the
Bureau.

There is also a broader limitation worth thinking about: you've looked at rainfall on its own,
without considering anything else that might also matter. What would you need to do to separate
rainfall's own effect from other factors that might be driving the same outcome? (This is what
the next session covers - once you've worked through it, you can come back to this dataset and
see what changes.)
